---
title: "Analysis of ITT's from AMPK KD Study"
author: "Katherine Kistler, Cody Cousineau, and Dave Bridges"
date: "July 5, 2019"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: yes
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

# Purpose

# Experimental Details

Injected 0.75 mU/g of lean body mass to animals fasted for 6h after 2 weeks of KD/CD (or four weeks after the injection)

# Raw Data

```{r data-input}
library(readxl) #loads the readr package

data.filename <- 'ITT 19-06-19.xlsx' #make this a separate line, you can use any variable you want
not.ko <- c(8830,8831,8715,8716,8717,8718,8719,8786,8346,8825,8826,8785,8787,8558)
exp.data <- read_excel(data.filename) %>% 
    filter(!(Mouse %in% not.ko)) #removed mice that are not ko

mapping.filename <- 'mapping.csv'
#this loads whatever the file is into a dataframe called exp.data if it exists
exp.data <- read_excel(data.filename, sheet='Sheet2') %>%
  select(Group:`120`) %>%
  filter(!(Group %in% c("Average","SE",NA))) %>% # removed calculated averages and errors
  separate(Group,
           into=c("Sex","Diet","Injection"), # broke group down into sex, diet and injection 
           sep=" ") %>%
  tibble::rowid_to_column("ID") %>% #index an ID column
  mutate(ID = as.factor(ID)) %>%
  mutate(Injection = relevel(as.factor(Injection), ref="GFP")) 

data.long <-
  exp.data %>%
  group_by(Sex,Diet,Injection) %>%
  gather(key=Time,value=Glucose, -Sex, -Diet, -Injection, -ID,-Mouse) %>%
  mutate(Time = as.integer(Time),
         Glucose = as.integer(Glucose))

summary.data <-
  data.long %>%
  group_by(Sex,Diet,Injection, Time) %>%
  mutate(Mouse = as.factor(Mouse)) %>%
  summarize_if(is.numeric, .funs=funs(mean=mean(., na.rm = TRUE),
                                      se=se))
```

These data can be found in **`r getwd()`** in a file named **`r data.filename`** and **`r mapping.filename`**.  This script was most recently updated on **`r date()`**.

# Number of Mice

```{r number-of-mice}
exp.data %>%
  group_by(Sex,Diet,Injection) %>%
  distinct(ID, .keep_all = T) %>%
  count %>%
  kable(caption="Animals in each group of this cohort")
```

# Analysis

```{r itt-points-all, fig.cap="Dotplot of ITT, all groups"}
library(ggplot2)

ggplot(data.long, aes(y=Glucose,
                      x=Time,
                      col=Injection)) +
  geom_point() +
  geom_smooth() +
  facet_grid(Sex~Diet) +
  labs(y="Blood Glucose (mg/dL)",
       x="Insulin (min)")
```


```{r itt-line-all, fig.cap="Lineplot of ITT, all mice"}
data.long %>%
  group_by(Sex,Diet,Injection,Time) %>%
  summarize(Mean = mean(Glucose,na.rm=T),
            Error = se(Glucose)) %>%
  ggplot(aes(y=Mean,
             x=Time,
             col=Injection)) +
  geom_line() +
  geom_errorbar(aes(ymin=Mean-Error,
             ymax=Mean+Error)) +
  expand_limits(y=0) +
  facet_grid(Sex~Diet) +
  labs(y="Blood Glucose (mg/dL)",
       x="Insulin (min)")
```

```{r itt-line-males, fig.cap="Lineplot of ITT, male mice only"}
data.long %>%
  filter(Sex=="M") %>%
  group_by(Sex,Diet,Injection,Time) %>%
  summarize(Mean = mean(Glucose,na.rm=T),
            Error = se(Glucose)) %>%
  ggplot(aes(y=Mean,
             x=Time,
             ymin=Mean-Error,
             ymax=Mean+Error,
             col=Injection)) +
  geom_line() +
  geom_errorbar() +
  expand_limits(y=0) +
  scale_color_brewer(palette = "Set2",
                     name="", 
                     labels=c("AAV-Tbg-GFP",
                              "AAV-Tbg-Cre")) +
  facet_grid(~Diet) +
  labs(y="Blood Glucose (mg/dL)",
       x="Insulin (min)",
       title="Hepatic AMPK Knockout Mice") +
  theme_classic() +
  theme(
    text = element_text(size=24),
    legend.position = c(0.25,0.85),
    legend.key = element_blank(),
    legend.background = element_blank())
```

```{r itt-line-gfp, fig.cap="Lineplot of ITT, all GFP mice"}
data.long %>%
  filter(Injection=="GFP") %>%
  group_by(Sex,Diet,Time) %>%
  summarize(Mean = mean(Glucose,na.rm=T),
            Error = se(Glucose)) %>%
  ggplot(aes(y=Mean,
             x=Time,
             ymin=Mean-Error,
             ymax=Mean+Error,
             col=Diet)) +
  geom_line() +
  geom_errorbar() +
  expand_limits(y=0) +
  facet_grid(~Sex) +
  labs(y="Blood Glucose (mg/dL)",
       x="Insulin (min)",
       title="Control mice only, effect of KD")
```

### ITT Statistics

```{r itt-stats}
library(lme4)
library(broom)
library(lmerTest)
itt.lme <- lmer(Glucose~as.factor(Time)+Diet+Sex+Injection+(1|ID), data=data.long)
itt.lme.null <- lmer(Glucose~as.factor(Time)+Diet+Sex+(1|ID), data=data.long)

summary(itt.lme)

anova(itt.lme) %>% 
  tidy %>% 
  kable(caption="Type III Analysis of Variance Table with Satterthwaite's method for ITT mixed linear model.")

anova(itt.lme,itt.lme.null) %>% 
  tidy %>%
  kable(caption="Chi-squared test for effects of AAV injection on ITT.")
```

## Normalized ITT

```{r itt-norm-line-all, fig.cap="Dotplot of ITT, normalized to fasting glucose"}
data.long.norm <-
  data.long %>%
  group_by(ID) %>%
  mutate(Glucose.norm = Glucose/Glucose[Time==0]*100)

data.long.norm %>%
  group_by(Sex,Diet,Injection,Time) %>%
  summarize(Mean = mean(Glucose.norm,na.rm=T),
            Error = se(Glucose.norm)) %>%
  ggplot(aes(y=Mean,
             x=Time,
             ymin=Mean-Error,
             ymax=Mean+Error,
             col=Injection)) +
  geom_line() +
  geom_errorbar() +
  expand_limits(y=0) +
  facet_grid(Sex~Diet) +
  labs(y="Blood Glucose (% of Basal)",
       x="Insulin (min)")
```

```{r itt-norm-line-gfp, fig.cap="Lineplot of ITT from GFP injected mice only, normalized to fasting glucose"}
data.long.norm %>%
  filter(Injection=="GFP") %>%
  group_by(Sex,Diet,Injection,Time) %>%
  summarize(Mean = mean(Glucose.norm,na.rm=T),
            Error = se(Glucose.norm)) %>%
  ggplot(aes(y=Mean,
             x=Time,
             ymin=Mean-Error,
             ymax=Mean+Error,
             col=Diet)) +
  geom_line() +
  geom_errorbar() +
  expand_limits(y=0) +
  facet_grid(~Sex) +
  labs(y="Blood Glucose (% of Basal)",
       x="Insulin (min)")
```

## Fasting Glucose

```{r fasting-glucose-boxplot, fig.cap="Boxplot of glucose levels after 6h fast"}
ggplot(data.long.norm %>% filter(Time==0),
       aes(y=Glucose,x=Diet,
           col=Injection)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitterdodge()) +
  expand_limits(y=0) +
  facet_grid(.~Sex) +
  labs(y="Fasting Glucose (mg/dL)",
       title="Fasting Glucose")
```

```{r fasting-glucose-barplot, fig.cap="Barplot of glucose levels after 6h fast"}
data.long %>% 
  filter(Time==0) %>%
  group_by(Injection,Sex,Diet) %>%
  summarize(Mean = mean(Glucose,na.rm=T),
            Error = se(Glucose)) %>%
  ggplot(aes(y=Mean,
             x=Diet,
             fill=Injection)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(position=position_dodge(width=0.75),aes(group=Injection,
                                                        ymin=Mean-Error,
                                                        ymax=Mean+Error), width=0.5) +
  expand_limits(y=0) +
  facet_grid(.~Sex) +
  labs(y="Fasting Glucose (mg/dL)",
       title="Fasting Glucose")
```

### Fasting Glucose Statistics

```{r fasting-glucose-stats}
lm(Glucose ~ Diet + Sex + Injection, data = data.long %>% filter(Time==0)) %>%
  summary %>%
  tidy %>%
  kable(caption="Linear model for effects on fasting glucose levels.")
```

## Area Under the Curve

```{r auc-boxplot, fig.cap="Boxplot of the area under the curves for insulin tolerance tests."}
data.long %>%
  group_by(ID, Diet, Sex, Injection) %>%
  summarize(AUC = sum(Glucose,na.rm=T)) %>%
ggplot(aes(y=AUC,x=Diet,
           col=Injection)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitterdodge()) +
  expand_limits(y=0) +
  facet_grid(.~Sex) +
  labs(y="Glucose (mg/dL x minutes)",
       title="Area Under Curve")
```

```{r auc-barplot, fig.cap="Barplot of the area under the curves for insulin tolerance tests."}
data.long %>%
  group_by(ID, Diet, Sex, Injection) %>%
  summarize(AUC = sum(Glucose)) %>%
    group_by(Diet, Sex, Injection) %>%
  summarize(Mean = mean(AUC,na.rm=T),
            Error = se(AUC)) %>%
  ggplot(aes(y=Mean,
             x=Diet,
             fill=Injection)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(position=position_dodge(width=0.75),aes(group=Injection,
                                                        ymin=Mean-Error,
                                                        ymax=Mean+Error), width=0.5) +
  expand_limits(y=0) +
  facet_grid(.~Sex) +
  labs(y="Glucose (mg/dL x minutes)",
       title="Area Under Curve")
```

### AUC Statistics

```{r auc-stats}
lm(AUC ~ Diet + Sex + Injection, 
   data = data.long %>% 
     group_by(ID, Diet, Sex, Injection) %>%
     summarize(AUC = sum(Glucose))) %>% 
  summary %>%
  tidy %>%
  kable(caption="Linear model for effects on ITT area under curve.")
```

# Interpretation

A brief summary of what the interpretation of these results were

# Session Information

```{r session-information, echo=T}
sessionInfo()
```

