---
title: "Re-analysis of Marin-Carli et al sMEC Data"
author: "Dave Bridges"
date: "January 4, 2021"
bibliography: "../references.bib"
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: yes
    toc: yes
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=FALSE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)
library(tibble)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

# Purpose

To re-analyse cell populations from the Martin-Carli *et al*'s scRNAseq study of lactating mammary glands.  This work is described in @MartinCarli2020.  This follows the analysis flow suggested for Seurat 3.2 seen at https://satijalab.org/seurat/v3.2/pbmc3k_tutorial.html

# Data Input

Downloaded the data from GSE15889 and removed prefixes from filenames.

```{r data-entry}
library(Seurat)
martin.counts <- Read10X(data.dir = '.')
martin.dataset <- CreateSeuratObject(counts = martin.counts,
                              project = "sMEC", 
                             min.cells = 3, 
                             min.features = 200)
```

## Preprocessing

```{r pre-processing}
martin.dataset[["percent.mt"]] <- PercentageFeatureSet(martin.dataset, pattern = "^MT-")
VlnPlot(martin.dataset, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), 
        ncol = 3)

plot1 <- FeatureScatter(martin.dataset, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(martin.dataset, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```
## Normalizing Data 

Normalizes the feature expression for each cell by the total expression  x 1000, then log transforms

```{r normalize}
martin.dataset <- NormalizeData(martin.dataset, 
                                normalization.method = "LogNormalize",
                                scale.factor = 10000)
```

## Highly Variable Features

Features with high cell to cell variability (highly expressed in some cells but not others)

```{r variable-features}
martin.dataset <- FindVariableFeatures(martin.dataset, 
                                       election.method = "vst", 
                                       nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(martin.dataset), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(martin.dataset)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

## Scaling

Shifts expression so that mean expression across cells is 0, and variance is 1.  This reduces the impact of outliers on downstream analyses.  We did not regress out specific sources of heterogeneity like mitochondrial contamination or cell cycle stage.

```{r scaling}
all.genes <- rownames(martin.dataset)
martin.dataset <- ScaleData(martin.dataset, features = all.genes)
```

# Dimensionality Reductions

```{r dimensionality-pca}
martin.dataset <- RunPCA(martin.dataset, features = VariableFeatures(object = martin.dataset))

print(martin.dataset[["pca"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(martin.dataset, dims = 1:2, reduction = "pca")
DimPlot(martin.dataset, reduction = "pca")
DimHeatmap(martin.dataset, dims = 1:9, cells = 500, balanced = TRUE)
```

# Determination of Number of Clusters

Used the JackStraw procedure in @Macosko2015, sampling 1% of the data re-running the PCA and constructing a null distribution of feature scores, then repeating.  This identified 'significant' PCs.  We also did an elbow plot.

```{r cluster-numbers}
martin.dataset <- JackStraw(martin.dataset, num.replicate = 100)
martin.dataset <- ScoreJackStraw(martin.dataset, dims=1:20)
JackStrawPlot(martin.dataset, dims=1:20)
ElbowPlot(martin.dataset)

pcs.used <- 7
```

Based on this we decided to use `r pcs.used` PCs to cluster the cells.  

#  Clustering Cell Types

Seurat 3.2 uses a K-nearest neighbor approach  then tries to partition this into commmunities of cell types.  

## Identification and Assignment of Clusters

```{r neighbor-cluster-identification}
martin.dataset <- FindNeighbors(martin.dataset, dims = 1:pcs.used) #based on 7 clusters from the elbow plot
martin.dataset <- FindClusters(martin.dataset) #default resolution is 0.8, use above 1.0 if you want a larger number of communities 
```


## Non-Linear Dimensionality Reduction

Did both UMAP and t-SNE plots using `r pcs.used` clusters

### t-SNE Plots

```{r tsne-plot-initial, fig.cap="t-SNE plot of secreted MECs"}
martin.dataset <- RunTSNE(martin.dataset, dims = 1:pcs.used, label=TRUE)
DimPlot(martin.dataset, reduction = "tsne")
#VlnPlot(martin.dataset, features = "UMAP_2")
```

### UMAP

```{r umap-plot-initial, fig.cap="UMAP plot of secreted MECs"}
martin.dataset <- RunUMAP(martin.dataset, dims = 1:pcs.used)
DimPlot(martin.dataset, reduction = "umap", label=T)
```


```{r cluster-classification}
cluster.markers.all <- FindAllMarkers(martin.dataset, 
                                      only.pos = TRUE,
                                      min.pct = 0.5, 
                                      logfc.threshold = 0.5)

cluster.markers.all %>% 
  group_by(cluster) %>% 
  top_n(n = 3, wt = avg_logFC) %>%
  kable(caption="Cell Specific Markers (All)")
```

# Feature Analysis

## Annotation of clusters

Used for cell marker enrichment

```{r cluster-data}
cell.marker.site <- 'http://bio-bigdata.hrbmu.edu.cn/CellMarker/download/Human_cell_markers.txt'
cell_marker_data <- vroom::vroom(cell.marker.site)
library(clusterProfiler)
## instead of `cellName`, users can use other features (e.g. `cancerType`)
cell_markers <- vroom::vroom(cell.marker.site) %>%
   tidyr::unite("cellMarker", tissueType, cancerType, cellName, sep=", ") %>% 
   dplyr::select(cellMarker, geneSymbol) %>%
   dplyr::mutate(geneSymbol = strsplit(geneSymbol, ', '))
```

Used CellMarker for enrichment analyses `r cell.marker.site`.  See @Zhang2019a and http://bio-bigdata.hrbmu.edu.cn/CellMarker for details on this resource

### Cluster 5

```{r cluster-annotation-5, fig.cap='Markers of cluster 5'}
# find all markers distinguishing cluster 5 from other communities
cluster5.markers <- FindMarkers(martin.dataset, ident.1 =5, 
                                logfc.threshold = 0.5, 
                                min.pct=0.25,
                                test.use = "roc", 
                                only.pos = TRUE) #using ROC as classification power
FeaturePlot(martin.dataset, features = rownames(cluster5.markers)[1:6]) 
VlnPlot(martin.dataset, features = rownames(cluster5.markers)[1:6], pt.size=0)

geneList <- cluster5.markers[,'avg_logFC']
names(geneList) <- rownames(cluster5.markers)
geneList <- sort(geneList, decreasing=T)

cm5 <- GSEA(geneList, TERM2GENE = cell_markers)
kable(cm5)

```

### Cluster 9

```{r cluster-annotation-9, fig.cap='Markers of cluster 9'}
# find all markers distinguishing cluster 9 from other communities
cluster9.markers <- FindMarkers(martin.dataset, ident.1 =9, 
                                logfc.threshold = 0.5, 
                                min.pct=0.25,
                                test.use = "roc", 
                                only.pos = TRUE)

FeaturePlot(martin.dataset, features = rownames(cluster9.markers)[1:6]) 
VlnPlot(martin.dataset, features = rownames(cluster9.markers)[1:6], pt.size=0)

geneList <- cluster9.markers[,'avg_logFC']
names(geneList) <- rownames(cluster9.markers)
geneList <- sort(geneList, decreasing=T)

cm9 <- GSEA(geneList, TERM2GENE = cell_markers)
kable(cm9)
```

### Cluster 6

```{r cluster-annotation-6, fig.cap='Markers of cluster 6'}
# find all markers distinguishing cluster 96 from other communities
cluster6.markers <- FindMarkers(martin.dataset, ident.1 =6, 
                                logfc.threshold = 0.5, 
                                min.pct=0.25,
                                test.use = "roc", 
                                only.pos = TRUE)

FeaturePlot(martin.dataset, features = rownames(cluster6.markers)[1:6]) 
VlnPlot(martin.dataset, features = rownames(cluster6.markers)[1:6], pt.size=0)

geneList <- cluster6.markers[,'avg_logFC']
names(geneList) <- rownames(cluster6.markers)
geneList <- sort(geneList, decreasing=T)

cm6 <- GSEA(geneList, TERM2GENE = cell_markers)
kable(cm6)
```

### Cluster 12

```{r cluster-annotation-12, fig.cap='Markers of cluster 12'}
cluster12.markers <- FindMarkers(martin.dataset, ident.1 =12, 
                                logfc.threshold = 0.5, 
                                min.pct=0.25,
                                test.use = "roc", 
                                only.pos = TRUE)

FeaturePlot(martin.dataset, features = rownames(cluster12.markers)[1:6]) 
VlnPlot(martin.dataset, features = rownames(cluster12.markers)[1:6], pt.size=0)

geneList <- cluster12.markers[,'avg_logFC']
names(geneList) <- rownames(cluster12.markers)
geneList <- sort(geneList, decreasing=T)

cm12 <- GSEA(geneList, TERM2GENE = cell_markers)
kable(cm12)
```

### Cluster 10

```{r cluster-annotation-10, fig.cap='Markers of cluster 10'}
cluster10.markers <- FindMarkers(martin.dataset, ident.1 =10, 
                                logfc.threshold = 0.5, 
                                min.pct=0.25,
                                test.use = "roc", 
                                only.pos = TRUE)

FeaturePlot(martin.dataset, features = rownames(cluster10.markers)[1:6]) 
VlnPlot(martin.dataset, features = rownames(cluster10.markers)[1:6], pt.size=0)

geneList <- cluster10.markers[,'avg_logFC']
names(geneList) <- rownames(cluster10.markers)
geneList <- sort(geneList, decreasing=T)

cm10 <- GSEA(geneList, TERM2GENE = cell_markers)
kable(cm10)
```

### Other

```{r cluster-annotation-other, fig.cap='Markers of cluster 12'}
cluster.all.markers <- FindMarkers(martin.dataset, 
                                ident.1 =c(0,1,2,3,4,7,8,11),
                                ident.2 =c(5,9,6,12),
                                logfc.threshold = 0.5, 
                                min.pct=0.25,
                                test.use = "roc", 
                                only.pos = TRUE)

FeaturePlot(martin.dataset, features = rownames(cluster.all.markers)[1:6]) 
VlnPlot(martin.dataset, features = rownames(cluster.all.markers)[1:6], pt.size=0)

geneList <- cluster.all.markers[,'avg_logFC']
names(geneList) <- rownames(cluster.all.markers)
geneList <- sort(geneList, decreasing=T)

cm.all <- GSEA(geneList, TERM2GENE = cell_markers)
kable(cm.all)
```

## Summary of Clusters

```{r cluster-summary}
new.cluster.ids <- c(0,1,2,3,4,'T-Cells','Monocytes',7,8,'Monocytes/Dendritic Cells',10,11,12)
names(new.cluster.ids) <- levels(martin.dataset)
martin.dataset <- RenameIdents(martin.dataset, new.cluster.ids)
DimPlot(martin.dataset, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```
,
## Analysis of clusters

```{r feature-analysis}
genes.of.interest <- c("Wap","Csn2","Csn3","Oxtr","Acta2","Krt14","Fasn", "Acly","Fabp4","Slc27a6") %>% toupper
protein.secreted <- c("Wap","Csn2","Csn3","Ltf","Lalba") %>% toupper
hormone.sensing <- c('Esr1', 'Prlr', 'Pgr', 'S100a6','Cited1') %>% toupper
myoepithelial <- c('Oxtr', 'Acta2', 'Krt4','Krt14') %>% toupper
alveolar.differentiated <- c('Wap', 'Csn2', 'Glycam1','Lalba') %>% toupper
alveolar.progenitor <- c('Wap', 'Csn2', 'Glycam1','Lalba','Cd14','Kit') %>% toupper
basal.cells <- c('Krt4', 'Krt14', 'Pdpn', 'Etv5','Acta2') %>% toupper
milk.fat.globules <- c('Pas3','Sluc1','Btn1a1','Pas7','Muc1','Plin2') %>% toupper
carb.transporters <-c('Slc2a1','Slc35a2','Slc50a1') %>% toupper
fat.transporters <-c('Fabp3','Cd36','Agpat2','Agpat3') %>% toupper
t.cells <-c('CD3','CD4','CD8','CD7','CTLA4') %>% toupper
b.cells <-c('CD19','CD27','CD20','IgD','CD38','CD138','CD24') %>% toupper
interleukins <-c('IL6','TNFA','IL10') %>% toupper
mtor <-c('RPTOR','MTOR','LST8') %>% toupper
cluster.markers.all %>%
  filter(gene %in% genes.of.interest) %>%
  kable(caption="Clusters of genes of interest")

VlnPlot(martin.dataset, features = genes.of.interest)

p1 <- FeaturePlot(martin.dataset, features = genes.of.interest,
            combine=F,
            ncol=2)
library(ggplot2)
fix.sc <- scale_color_gradientn( colours = c('lightgrey', 'blue'),  limits = c(1, 8))
p2 <- lapply(p1, function (x) x + fix.sc)
CombinePlots(p2)
for(i in 1:length(p2)) {
  p2[[i]] <- p2[[i]] + NoLegend() + NoAxes()
}

cowplot::plot_grid(plotlist = p2)
FeaturePlot(martin.dataset, features = protein.secreted) 
FeaturePlot(martin.dataset, features = alveolar.progenitor)
FeaturePlot(martin.dataset, features = alveolar.differentiated)
FeaturePlot(martin.dataset, features = basal.cells)
FeaturePlot(martin.dataset, features = hormone.sensing)
FeaturePlot(martin.dataset, features = myoepithelial)
FeaturePlot(martin.dataset, features = milk.fat.globules)
FeaturePlot(martin.dataset, features = genes.of.interest)
FeaturePlot(martin.dataset, features = carb.transporters)
FeaturePlot(martin.dataset, features = fat.transporters)
FeaturePlot(martin.dataset, features = t.cells)
FeaturePlot(martin.dataset, features = b.cells)
FeaturePlot(martin.dataset, features = interleukins)
FeaturePlot(martin.dataset, features = mtor)


top10 <- cluster.markers.all %>% group_by(cluster) %>% top_n(n = 5, wt = avg_logFC)
DoHeatmap(martin.dataset, features = top10$gene) + NoLegend()
```



# Session Information

```{r session-information, echo=T}
sessionInfo()
```
