---
title: "Re-Analysis of GSE17190"
author: "Dave Bridges"
date: "May 11, 2020"
output:
  html_document:
    keep_md: yes
  pdf_document:
    keep_tex: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=FALSE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively
knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})
superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)
  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))
#load these packages, nearly always needed
library(tidyr)
library(dplyr)
library(broom)
# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

This is generated via GEO2R based on the Kivela *et al.* paper which did microarrays on gastroc muscles from LCR vs HCR rats.

```{r geo2r, results='asis'}
# Version info: R 3.2.3, Biobase 2.30.0, GEOquery 2.40.0, limma 3.26.8
# R scripts generated  Mon May 11 11:42:23 EDT 2020

################################################################
#   Differential expression analysis with limma
library(Biobase)
library(GEOquery)
library(limma)

# load series and platform data from GEO

gset <- getGEO("GSE17190", GSEMatrix =TRUE, AnnotGPL=TRUE)
if (length(gset) > 1) idx <- grep("GPL6101", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]

# make proper column names to match toptable 
fvarLabels(gset) <- make.names(fvarLabels(gset))

# group names for all samples
gsms <- "110000110011001100110011"
sml <- c()
for (i in 1:nchar(gsms)) { sml[i] <- substr(gsms,i,i) }

# log2 transform
ex <- exprs(gset)
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
          (qx[6]-qx[1] > 50 && qx[2] > 0) ||
          (qx[2] > 0 && qx[2] < 1 && qx[4] > 1 && qx[4] < 2)
if (LogC) { ex[which(ex <= 0)] <- NaN
  exprs(gset) <- log2(ex) }

# set up the data and proceed with analysis
sml <- paste("G", sml, sep="")    # set group names
fl <- as.factor(sml)
gset$description <- fl
design <- model.matrix(~ description + 0, gset)
colnames(design) <- levels(fl)
fit <- lmFit(gset, design)
cont.matrix <- makeContrasts(G1-G0, levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2, 0.01)
tT <- topTable(fit2, adjust="fdr", sort.by="B", number=250)

tT <- subset(tT, select=c("ID","adj.P.Val","P.Value","t","B","logFC","Gene.symbol","Gene.title"))
#write.table(tT, file=stdout(), row.names=F, sep="\t")


################################################################
#   Boxplot for selected GEO samples
library(Biobase)
library(GEOquery)

# load series and platform data from GEO

gset <- getGEO("GSE17190", GSEMatrix =TRUE, getGPL=FALSE)
if (length(gset) > 1) idx <- grep("GPL6101", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]

# group names for all samples in a series
gsms <- "110000110011001100110011"
sml <- c()
for (i in 1:nchar(gsms)) { sml[i] <- substr(gsms,i,i) }
sml <- paste("G", sml, sep="")  # set group names

# order samples by group
ex <- exprs(gset)[ , order(sml)]
sml <- sml[order(sml)]
fl <- as.factor(sml)
labels <- c("LCR","HCR")

# set parameters and draw the plot
palette(c("#dfeaf4","#f4dfdf", "#AABBCC"))
dev.new(width=4+dim(gset)[[2]]/5, height=6)
par(mar=c(2+round(max(nchar(sampleNames(gset)))/2),4,2,1))
title <- paste ("GSE17190", '/', annotation(gset), " selected samples", sep ='')
boxplot(ex, boxwex=0.6, notch=T, main=title, outline=FALSE, las=2, col=fl)
legend("topleft", labels, fill=palette(), bty="n")



#write to output file
output_file <- 'GSE1790 Analysis.csv'
write.fit(fit2, file=output_file, adjust='BH')

hcr.results <- droplevels(topTable(fit2, n=Inf, adjust.method="BH")) %>%
  select(Gene.symbol,logFC,P.Value,adj.P.Val) %>%
  distinct(Gene.symbol, .keep_all=T)
#annotate the probes
sig.hcr.results <- hcr.results %>% filter(adj.P.Val<0.05)
sig.hcr.genes <- sig.hcr.results$Gene.symbol

mTsc.datafile <- '../RNAseq/data/processed/Binary DESeq Results.csv'
mtsc.data <- read.csv(mTsc.datafile)
sig.mtsc.data <- droplevels(subset(mtsc.data, padj<0.05))

sig.mtsc.genes <- sig.mtsc.data$external_gene_name
sig.overlap <- intersect(sig.hcr.genes, sig.mtsc.genes)

fisher.table <-
  matrix(c(dim(sig.mtsc.data)[1],
           length(sig.overlap), 
           length(unique(fit2$genes$Gene.symbol)),
           length(unique(sig.hcr.genes))),
       nrow = 2,
       dimnames = list(Genes = c("Tested", "Sig"),
                       Significance = c("Overlap", "HCR")))
```

Duvel et al identified `r length(unique(sig.hcr.genes))` significantly differentially expressed genes in their analysis out of a total of `r length(unique(fit2$genes$Gene.symbol))` genes assessed.  Of these differentially expressed genes. `r length(sig.overlap)` genes overlapped with our `r dim(sig.mtsc.data)[1]` significantly different genes from *Tsc1* knockout mice quadriceps.  This is not a significant level of overlap (p=`r fisher.test(fisher.table)$p.value`).

```{r gene-overlap}
combined.genes <-
  mtsc.data %>%
  select(external_gene_name,log2FoldChange,pvalue,padj) %>%
  left_join(hcr.results, by=c('external_gene_name'='Gene.symbol'))

library(ggplot2)
combined.genes %>%
  ggplot(aes(y=log2FoldChange,x=logFC)) +
  geom_point(size=0.1, alpha=0.1) +
  labs(y="TSC Fold Change",x="HCR Fold Change") +
  geom_abline()


interesting.genes <- c('Slc16a1','Oxct1','Bdh1','Gdf15','Fabp3','Cd36')
combined.genes %>%
  filter(external_gene_name %in% interesting.genes)
```

```{r hcr-mtsc-venn}
require(venneuler)
v.diseases <- venneuler(c("HCR vs LCR"=length(sig.hcr.genes), 
                 "Tsc1 Knockout Muscles"=length(sig.mtsc.genes),
                 "HCR vs LCR&Tsc1 Knockout Muscles"=length(intersect(sig.hcr.genes, sig.mtsc.genes))))

plot(v.diseases, main="TSC-Dependent Transcriptional Changes")

# library(Vennerable)
# v.list <- list(`Tsc2 Knockout MEFs` = sig.duvel.genes, `Tsc1 Knockout Muscles` = sig.mtsc.genes)
# v.data <- Venn(v.list)
# plot(v.data)
```

```{r session-info}
sessionInfo()
```