---
title: "Re-Analysis of GSE20104 - ATF4 Overexpression Muscles"
author: "Dave Bridges"
date: "February 7, 2022"
output:
  html_document:
    keep_md: yes
  pdf_document:
    keep_tex: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively
knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})
superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)
  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))
#load these packages, nearly always needed
library(tidyr)
library(dplyr)
library(broom)
# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

This is generated via GEO2R.  Ignored fed and fasted mice

```{r GSE20104-geo2r, results='asis'}
# Version info: R 3.2.3, Biobase 2.30.0, GEOquery 2.40.0, limma 3.26.8
################################################################
#   Differential expression analysis with limma
library(GEOquery)
library(limma)
library(umap)

# load series and platform data from GEO

gset <- getGEO("GSE20104", GSEMatrix =TRUE, AnnotGPL=FALSE)
if (length(gset) > 1) idx <- grep("GPL6096", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]

# make proper column names to match toptable 
fvarLabels(gset) <- make.names(fvarLabels(gset))

# group membership for all samples
gsms <- "01010101XXXXXXXX"
sml <- strsplit(gsms, split="")[[1]]

# filter out excluded samples (marked as "X")
sel <- which(sml != "X")
sml <- sml[sel]
gset <- gset[ ,sel]

# log2 transformation
ex <- exprs(gset)
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
          (qx[6]-qx[1] > 50 && qx[2] > 0)
if (LogC) { ex[which(ex <= 0)] <- NaN
  exprs(gset) <- log2(ex) }

# assign samples to groups and set up design matrix
gs <- factor(sml)
groups <- make.names(c("Control","ATF4 OE"))
levels(gs) <- groups
gset$group <- gs
gs <- relevel(gs, ref="Control")
design <- model.matrix(~group + 0, gset)
colnames(design) <- levels(gs)

fit <- lmFit(gset, design)  # fit linear model

# set up contrasts of interest and recalculate model coefficients
cts <- paste(groups[2], groups[1], sep="-")
cont.matrix <- makeContrasts(contrasts=cts, levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)

# compute statistics and table of top significant genes
fit2 <- eBayes(fit2, 0.01)
tT <- topTable(fit2, adjust="fdr", sort.by="B", number=250)

tT <- subset(tT, select=c("ID","adj.P.Val","P.Value","t","B","logFC","GB_LIST","SPOT_ID","RANGE_GB","RANGE_STRAND","RANGE_START"))
#write.table(tT, file=stdout(), row.names=F, sep="\t")

# Visualize and quality control test results.
# Build histogram of P-values for all genes. Normal test
# assumption is that most genes are not differentially expressed.
tT2 <- topTable(fit2, adjust="fdr", sort.by="B", number=Inf) %>%
  separate(gene_assignment, into=c('Genbank','Gene.symbol','Gene Name','Other'),
           sep=" // ",
           extra="merge")
hist(tT2$adj.P.Val, col = "grey", border = "white", xlab = "P-adj",
  ylab = "Number of genes", main = "P-adj value distribution")

# summarize test results as "up", "down" or "not expressed"
dT <- decideTests(fit2, adjust.method="fdr", p.value=0.05)

# Venn diagram of results
vennDiagram(dT, circle.col=palette())

# create Q-Q plot for t-statistic
t.good <- which(!is.na(fit2$F)) # filter out bad probes
qqt(fit2$t[t.good], fit2$df.total[t.good], main="Moderated t statistic")

# volcano plot (log P-value vs log fold change)
colnames(fit2) # list contrast names
ct <- 1        # choose contrast of interest
volcanoplot(fit2, coef=ct, main=colnames(fit2)[ct], pch=20,
  highlight=length(which(dT[,ct]!=0)), names=rep('+', nrow(fit2)))

# MD plot (log fold change vs mean log expression)
# highlight statistically significant (p-adj < 0.05) probes
plotMD(fit2, column=ct, status=dT[,ct], legend=F, pch=20, cex=1)
abline(h=0)

################################################################
# General expression data analysis
ex <- exprs(gset)

# box-and-whisker plot
ord <- order(gs)  # order samples by group
palette(c("#1B9E77", "#7570B3", "#E7298A", "#E6AB02", "#D95F02",
          "#66A61E", "#A6761D", "#B32424", "#B324B3", "#666666"))
par(mar=c(7,4,2,1))
title <- paste ("GSE20104", "/", annotation(gset), sep ="")
boxplot(ex[,ord], boxwex=0.6, notch=T, main=title, outline=FALSE, las=2, col=gs[ord])
legend("topleft", groups, fill=palette(), bty="n")

# expression value distribution
par(mar=c(4,4,2,1))
title <- paste ("GSE20104", "/", annotation(gset), " value distribution", sep ="")
plotDensities(ex, group=gs, main=title, legend ="topright")

# UMAP plot (dimensionality reduction)
ex <- na.omit(ex) # eliminate rows with NAs
ex <- ex[!duplicated(ex), ]  # remove duplicates
ump <- umap(t(ex), n_neighbors = 4, random_state = 123)
par(mar=c(3,3,2,6), xpd=TRUE)
plot(ump$layout, main="UMAP plot, nbrs=4", xlab="", ylab="", col=gs, pch=20, cex=1.5)
legend("topright", inset=c(-0.15,0), legend=levels(gs), pch=20,
col=1:nlevels(gs), title="Group", pt.cex=1.5)
library("maptools")  # point labels without overlaps
pointLabel(ump$layout, labels = rownames(ump$layout), method="SANN", cex=0.6)

# mean-variance trend, helps to see if precision weights are needed
plotSA(fit2, main="Mean variance trend, GSE20104")


#write to output file
output_file <- 'GSE20104 Analysis.csv'
write.fit(fit2, file=output_file, adjust='BH')

atf4.oe.results <- droplevels(topTable(fit2, n=Inf, adjust.method="BH")) %>%
  separate(gene_assignment, into=c('Genbank','Gene.symbol','Gene Name','Other'),
           sep=" // ",
           extra="merge") %>%
  select(Gene.symbol,logFC,P.Value,adj.P.Val) %>%
  distinct(Gene.symbol, .keep_all=T)
#annotate the probes
sig.atf4.oe.results <- atf4.oe.results %>% filter(P.Value<0.05)
sig.atf4.oe.genes <- sig.atf4.oe.results$Gene.symbol#used nominal p value, nothing sig after FDR

sig.atf4.oe.results.up <- atf4.oe.results %>% filter(P.Value<0.05&logFC>0)#used nominal p value, nothing sig after FDR
sig.atf4.oe.genes.up <- sig.atf4.oe.results.up$Gene.symbol

sig.atf4.oe.results.down <- atf4.oe.results %>% filter(P.Value<0.05&logFC<0)#used nominal p value, nothing sig after FDR
sig.atf4.oe.genes.down <- sig.atf4.oe.results.down$Gene.symbol

mTsc.datafile <- '../RNAseq/Muscle Tsc1 Knockout/data/processed/Binary DESeq Results.csv'
mtsc.data <- read.csv(mTsc.datafile)

sig.mtsc.data <- droplevels(subset(mtsc.data, padj<0.05))
sig.mtsc.data.up <- droplevels(subset(mtsc.data, padj<0.05&log2FoldChange>0))
sig.mtsc.data.down <- droplevels(subset(mtsc.data, padj<0.05&log2FoldChange<0))

sig.mtsc.genes <- sig.mtsc.data$external_gene_name
sig.mtsc.genes.up <- sig.mtsc.data.up$external_gene_name
sig.mtsc.genes.down <- sig.mtsc.data.down$external_gene_name

sig.overlap <- intersect(sig.atf4.oe.genes, sig.mtsc.genes)
sig.overlap.up <- intersect(sig.atf4.oe.genes.up, sig.mtsc.genes.up)
sig.overlap.down <- intersect(sig.atf4.oe.genes.down, sig.mtsc.genes.down)

fisher.table <-
  matrix(c(length(unique(sig.atf4.oe.genes)),
           length(unique(atf4.oe.results$Gene.symbol)), 
           length(sig.overlap),
           dim(sig.mtsc.data)[1]),
       nrow = 2,
       dimnames = list(KD.Sig = c("Sig", "Total"),
                       mTSC.Sig = c("ATF4", "mTSC")))

kable(fisher.table, caption="Contingency table for comparason of ATF4 and mTSC, all genes")

fisher.table.up <-
  matrix(c(length(unique(sig.atf4.oe.genes.up)),
           length(unique(atf4.oe.results$Gene.symbol)), 
           length(sig.overlap.up),
           dim(sig.mtsc.data)[1]),
       nrow = 2,
       dimnames = list(KD.Sig = c("Sig", "Total"),
                       mTSC.Sig = c("ATF4", "mTSC")))

kable(fisher.table.up, caption="Contingency table for comparason of ATF4 and mTSC, up-regulated genes")

fisher.table.down <-
  matrix(c(length(unique(sig.atf4.oe.genes.down)),
           length(unique(atf4.oe.results$Gene.symbol)), 
           length(sig.overlap.down),
           dim(sig.mtsc.data)[1]),
       nrow = 2,
       dimnames = list(KD.Sig = c("Sig", "Total"),
                       mTSC.Sig = c("ATF4", "mTSC")))

kable(fisher.table.down, caption="Contingency table for comparason of ATF4 and mTSC, down-regulated genes")
```

This identified `r length(unique(sig.atf4.oe.genes))` significantly differentially expressed genes in their analysis out of a total of `r length(unique(atf4.oe.results$Gene.symbol))` genes assessed.  Of these differentially expressed genes. `r length(sig.overlap)` genes overlapped with our `r dim(sig.mtsc.data)[1]` significantly different genes from *Tsc1* knockout mice quadriceps.  This is an insignificant level of overlap (`r fisher.test(fisher.table)$estimate` fold enrichment; p=`r fisher.test(fisher.table)$p.value`).

Focusing specifically on directionality, this is a significant level of overlap for up (`r fisher.test(fisher.table.up)$estimate` times more likely; p=`r fisher.test(fisher.table.up)$p.value`) and dowregulated genes (`r fisher.test(fisher.table.down)$estimate` times more likely; p=`r fisher.test(fisher.table.down)$p.value`).  

For upregulation this represents `r length(unique(sig.atf4.oe.genes.up))` ATF4 dependent genes out of `r length(unique(atf4.oe.results$Gene.symbol))` genes assessed.  Of these differentially expressed genes. `r length(sig.overlap.up)` genes overlapped with our `r dim(sig.mtsc.data)[1]` significantly upregulated genes from *Tsc1* knockout mice quadriceps.  

For downregulation this represents `r length(unique(sig.atf4.oe.genes.down))` ATF4 dependent genes out of `r length(unique(atf4.oe.results$Gene.symbol))` genes assessed.  Of these differentially expressed genes. `r length(sig.overlap.down)` genes overlapped with our `r dim(sig.mtsc.data)[1]` significantly downregulated genes from *Tsc1* knockout mice quadriceps.  

```{r atf4-mtsc-gene-overlap}
combined.genes <-
  mtsc.data %>%
  select(external_gene_name,log2FoldChange,pvalue,padj) %>%
  left_join(atf4.oe.results, by=c('external_gene_name'='Gene.symbol')) %>%
  rename("log2FC_TSC"="log2FoldChange",
         "pval_TSC"='pvalue',
         "padj_TSC"='padj',
         "log2FC_KD"="logFC",
         'pval_KD'="P.Value",
         'padj_KD'='adj.P.Val')

with(combined.genes, lm(log2FC_TSC~log2FC_KD))  %>% tidy %>% kable(caption="Linear model for association between ATF4 fold change and TSC fold change")


library(ggplot2)
combined.genes %>%
  ggplot(aes(y=log2FC_TSC,x=log2FC_KD)) +
  geom_point(size=0.1, alpha=0.1) +
  labs(y="TSC Fold Change",x="KD Fold Change") +
  geom_smooth(se=T)

interesting.genes <- c('Sln','Gdf15') #
combined.genes %>%
  filter(external_gene_name %in% interesting.genes) %>%
  kable(caption="Selected genes")

combined.genes %>%
  filter(external_gene_name %in% interesting.genes) %>%
  select(external_gene_name, log2FC_KD, log2FC_TSC) %>%
  group_by(external_gene_name) %>%
  pivot_longer(names_to = 'Experiment', values_to = 'Log2FC', cols=log2FC_KD:log2FC_TSC) %>%
  ggplot(aes(y=Log2FC,x=external_gene_name,
             fill=Experiment)) +
    geom_bar(stat='identity', position='dodge')
```

```{r atf4-mtsc-venn}
require(venneuler)
v.diseases <- venneuler(c("ATF4"=length(sig.atf4.oe.genes), 
                 "Tsc1 Knockout Muscles"=length(sig.mtsc.genes),
                 "ATF4&Tsc1 Knockout Muscles"=length(intersect(sig.atf4.oe.genes, sig.mtsc.genes))))

v.diseases.up <- venneuler(c("ATF4"=length(sig.atf4.oe.genes.up), 
                 "Tsc1 Knockout Muscles"=length(sig.mtsc.genes.up),
                 "ATF4&Tsc1 Knockout Muscles"=length(intersect(sig.atf4.oe.genes.up, sig.mtsc.genes.up))))
v.diseases.down <- venneuler(c("ATF4"=length(sig.atf4.oe.genes.down), 
                 "Tsc1 Knockout Muscles"=length(sig.mtsc.genes.down),
                 "ATF4&Tsc1 Knockout Muscles"=length(intersect(sig.atf4.oe.genes.down, sig.mtsc.genes.down))))

plot(v.diseases, main="TSC-Dependent Transcriptional Changes")
plot(v.diseases.up, main="TSC-Dependent and ATF4 Upregulation")
plot(v.diseases, main="TSC-Dependent and ATF4 Downregulation")

# library(Vennerable)
# v.list <- list(`Tsc2 Knockout MEFs` = sig.duvel.genes, `Tsc1 Knockout Muscles` = sig.mtsc.genes)
# v.data <- Venn(v.list)
# plot(v.data)
```

# Pathway Analysis

```{r atf4-pathway-analysis}
library(clusterProfiler)
library(org.Mm.eg.db)
db.db.logfc <- 
  tT2 %>% 
  arrange(-logFC) %>%
  dplyr::select(logFC, Gene.symbol) 

db.db.list <- as.numeric(db.db.logfc$logFC)
names(db.db.list) <- db.db.logfc$Gene.symbol
gsea.bp <- gseGO(db.db.list, 
             ont ="BP", 
             keyType = "SYMBOL", 
             OrgDb = org.Mm.eg.db)
library(enrichplot)
upsetplot(gsea.bp)
dotplot(gsea.bp, orderBy="NES")
gsea.bp <- pairwise_termsim(gsea.bp, method="JC")
emapplot(gsea.bp,
         cex_label_category=0.5,
         showCategory = 50,
         color="NES",
         min_edge=0.1)

gsea.bp %>% as.data.frame %>% dplyr::select(NES,pvalue,p.adjust,Description, core_enrichment) %>% arrange(-NES) %>% filter(p.adjust<0.05) %>% kable(caption="Significant GO-BP pathways")
```

## Empirical Muscle ATF4 Gene Set

```{r gene-set-atf4}
list(ATF4q0.25 = tT2 %>% filter(adj.P.Val <0.25) %>% pull(Gene.symbol),
     ATF4p0.001 = tT2 %>% filter(P.Value <0.001) %>% pull(Gene.symbol),
     ATF4p0.01 = tT2 %>% filter(P.Value <0.01) %>% pull(Gene.symbol),
     ATF4p0.05 = tT2 %>% filter(P.Value <0.05) %>% pull(Gene.symbol)) -> atf4.gene.lists
library(fgsea)

mtsc.ranks <- pull(mtsc.data, log2FoldChange)
names(mtsc.ranks) <- mtsc.data$external_gene_name

mtsc.atf4.pathways <- fgsea(pathways = atf4.gene.lists, 
                  stats    = sort(mtsc.ranks,decreasing=T),
                  minSize  = 1,
                  maxSize  = 500)

mtsc.atf4.pathways %>% 
  arrange(-NES) %>%
  kable(caption="GSEA results for mTSC differential expression results compared to empirically determined ATF4 dependent changes in muscle")
plotEnrichment(atf4.gene.lists[["ATF4p0.001"]],
               !(is.na(sort(mtsc.ranks, decreasing=T))))

plotGseaTable(atf4.gene.lists, !(is.na(sort(mtsc.ranks,decreasing=T))), mtsc.atf4.pathways, 
              gseaParam=0.5)

plotEnrichment(atf4.gene.lists[["ATF4p0.01"]],
               !(is.na(sort(mtsc.ranks, decreasing=T)))) +
  labs(title="ATF4 Targets in Muscle Tsc1 Knockouts") +
  theme_classic() +
  theme(text=element_text(size=16))
  
```

## Pathway Analysis of Dual TSC/ATF4 Genes

The co-regulated genes in TSC and ATF4 muscles were:

*Upregulated*: `r sig.overlap.up %>% paste(collapse=", ")`

*Downregulated*:`r sig.overlap.down %>% paste(collapse=", ")`


```{r atf4-mtsc-pathway-analysis}

enrich.bp <- enrichGO(c(sig.overlap.up,sig.overlap.down), 
             ont ="BP", 
             keyType = "SYMBOL", 
             OrgDb = org.Mm.eg.db)
library(enrichplot)
#upsetplot(enrich.bp)
#dotplot(enrich.bp, orderBy="NES")
#enrich.bp <- pairwise_termsim(enrich.bp, method="JC")
# emapplot(enrich.bp,
#          cex_label_category=0.5,
#          showCategory = 50,
#          color="NES",
#          min_edge=0.1)

enrich.bp %>% as.data.frame %>% dplyr::select(GeneRatio,pvalue,p.adjust,Description) %>%  filter(p.adjust<0.05) %>% kable(caption="Significant GO-BP pathways for co-regulated genes")
```

# Specific Genes

## Sarcolipin

```{r atf4-sln-barplot}

sln.id <- tT2 %>% filter(Gene.symbol=='Sln') %>% rownames()
sln.exprs <- exprs(gset)[sln.id,]

sln.data <- data.frame(sln.exprs, gs)

library(forcats)
sln.data %>%
  group_by(gs) %>%
  summarize(avg=mean(2^sln.exprs),
            error=se(2^sln.exprs)) %>%
  mutate(norm.mean = avg/avg[1],
         norm.error = error/avg[1]) %>%
  ggplot(aes(x=gs,
             y=norm.mean, 
             ymin=norm.mean-norm.error,
             ymax=norm.mean+norm.error)) +
  geom_bar(stat='identity', position='dodge') +
  geom_errorbar(width=0.5) +
  scale_x_discrete(labels=c("Control" = "Empty Vector", "ATF4.OE" = "ATF4")) +
  theme_classic() +
  theme(text=element_text(size=20)) +
  labs(y='Relative mRNA Expression',
       x='',
       title='Effects ATF4 Overexpression on Sarcolipin')
```

## Ketolytic Genes

```{atf4-ketolytic-barplot}
genes.to.test <- c('Slc16a1','Bdh1','Hmgcs2')
kt.id <- tT2 %>% filter(Gene.symbol%in%genes.to.test) %>% rownames()
kt.exprs <- exprs(gset)[kt.id,] 
rownames(kt.exprs) <- genes.to.test

library(tibble)
kt.data <- as.data.frame(kt.exprs) %>%
  rownames_to_column(var="Gene") %>%
  pivot_longer(-Gene, names_to='Sample') %>%
  mutate(Treatment=rep(as.character(gs),3))


library(forcats)
kt.data %>%
  group_by(Gene,Treatment) %>%
  summarize(avg=mean(2^sln.exprs),
            error=se(2^sln.exprs),
            .groups='keep') %>% #not working here
  mutate(norm.mean = avg/avg[Treatment=='Control'],
         norm.error = error/avg[Treatment=='Control']) %>%
  ggplot(aes(x=Treatment,
             y=avg
             ymin=norm.mean-norm.error,
             ymax=norm.mean+norm.error)) +
  geom_bar(stat='identity', position='dodge') +
  geom_errorbar(width=0.5) +
  scale_x_discrete(labels=c("Control" = "Empty Vector", "ATF4.OE" = "ATF4")) +
  theme_classic() +
  theme(text=element_text(size=16)) +
  labs(y='Relative mRNA Expression',
       x='',
       title='Effects ATF4 Overexpression on Ketolysis Genes')
```

# Session Information

```{r session-info}
sessionInfo()
```