---
title: "Body Composition of Cohort 3 AJ Mice on KD"
author: "Cody Cousineau and Dave Bridges"
date: "June 11, 2019"
output:
  html_document:
    highlight: tango
    keep_md: yes
    toc: yes
    fig_caption: yes
  pdf_document:
    highlight: tango
    keep_tex: yes
    toc: yes
    fig_caption: yes
---


```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
         length = length, ...)


se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)
library(readr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

This script can be found in `r getwd()` and was most recently run on `r date()`.


# Data Entry

```{r data-entry}

col.types <- cols(
  age = col_double(),
  animal.id = col_factor(level=NULL),
  MouseID = col_factor(level=NULL),
  Sex = col_factor(level=NULL),
  Genotype = col_character(),
  values = col_double(),
  assay = col_factor(level=NULL),
  experiment.date = col_date(format = ""),
  Cohort = col_factor(levels=NULL),
  Mass = col_double()
)

filename <- 'Combined Body Composition Data.csv'
sick.mice <- c(4099,	8067)
exp.data <- read_csv(filename, col_types = col.types) %>%
  filter(diet.time<4) %>% #remove data on fourth week
  filter(!(MouseID %in% sick.mice)) # removed sick mice
```

This script pulled in a total of `r dim(exp.data)[1]` observations.  This includes the following number of animals in each treatment group. 

# Enrolled Animals

This is for animals where wer have any body composition data.

```{r enrollees}
exp.data %>%
  group_by(Sex,Diet) %>%
  distinct(animal.id, .keep_all = T) %>%
  count %>%
  kable(caption="Animals in each group of this cohort")

exp.data %>%
  group_by(Sex,Diet) %>%
  distinct(animal.id, .keep_all = T) %>%
  arrange(Sex,Diet) %>%
  select(MouseID,Sex,Diet) %>%
  kable(caption="Animals in each group of this cohort")
```

# Body Weight

```{r body-weight-scatterplot, fig.cap="Scatter Plot of Body Weights"}
library(ggplot2)

exp.data %>%
  filter(assay=='Body Weight') %>%
  ggplot(aes(y=Mass,x=diet.time,col=Diet)) +
  geom_point() +
    facet_grid(.~Sex) +
  stat_smooth() +
  labs(title="Body Weight",
       x="Time on Diet (Weeks)",
       y="Body Weight (g)")
```

```{r body-weight-individual, fig.cap="Line Plot of Individual Body Weights"}
exp.data %>%
  filter(assay=='Body Weight') %>%
  ggplot(aes(y=Mass,x=diet.time,col=Diet,group=animal.id)) +
  geom_line() +
  geom_point() +
  facet_grid(.~Sex) +
  labs(title="Body Weight",
       subtitle="Individual Trajectories",
       x="Time on Diet (Weeks)",
       y="Body Weight (g)") 
```

```{r body-weight-lineplot, fig.cap="Line Plot of Body Weights"}
exp.data %>%
  filter(assay=='Body Weight') %>%
  group_by(Diet,Sex,diet.time) %>%
  summarize(Average = mean(Mass),
            SE = se(Mass)) %>%
  ggplot(aes(y=Average,
             ymin=Average-SE,
             ymax=Average+SE,
             x=diet.time,col=Diet)) +
  geom_errorbar(width=0.5) +
  geom_line() +
    facet_grid(.~Sex) +
  labs(title="Body Weight",
       x="Time on Diet (Weeks)",
       y="Body Weight (g)")

exp.data %>%
  filter(assay=='Body Weight') %>%
  group_by(Diet,Sex,diet.time) %>%
  summarize(Average = mean(Mass),
            SE = se(Mass)) %>%
  ggplot(aes(y=Average,
             ymin=Average-SE,
             ymax=Average+SE,
             x=diet.time,col=Diet)) +
  geom_errorbar(width=0.5) +
  geom_line() +
    facet_grid(.~Sex) +
  labs(title="",
       x="Time on Diet (Weeks)",
       y="Body Weight (g)") +
  scale_color_grey() +
  theme_classic() +
    theme(text=element_text(size=18),
          legend.position = c(0.75,0.75))
```

## Body Weight Statistics

```{r weight-stats}
library(lme4)
library(lmerTest)
male.lme.weight.null <- lmer(Mass ~ diet.time + (1|animal.id), data=filter(exp.data, assay=='Body Weight'&diet.time>=0&Sex=="M"))
male.lme.weight <- lmer(Mass ~ diet.time*Diet + (1|animal.id), data=filter(exp.data, assay=='Body Weight'&diet.time>=0&Sex=="M"))
anova(male.lme.weight,male.lme.weight.null) %>% kable(caption="Male weight stats")

female.lme.weight.null <- lmer(Mass ~ diet.time + (1|animal.id), data=filter(exp.data, assay=='Body Weight'&diet.time>=0&Sex=="F"))
female.lme.weight <- lmer(Mass ~ diet.time*Diet + (1|animal.id), data=filter(exp.data, assay=='Body Weight'&diet.time>=0&Sex=="F"))
anova(female.lme.weight,female.lme.weight.null) %>% kable(caption="Female weight stats")

lmer(Mass ~ diet.time*Diet*Sex + (1|animal.id), data=filter(exp.data, assay=='Body Weight'&diet.time>=0)) %>% summary %>% coef%>% kable(caption="Interaction of sex on body weight")
```

```{r body-weight-lineplot-change, fig.cap="Line Plot of Change in Body Weights"}
exp.data %>%
  filter(assay=='Body Weight') %>%
  group_by(Diet,Sex,diet.time) %>%
  summarize(Average = mean(Mass.Change, na.rm=T),
            SE = se(Mass.Change)) %>%
  ggplot(aes(y=Average,
             ymin=Average-SE,
             ymax=Average+SE,
             x=diet.time,col=Diet)) +
  geom_errorbar(width=0.5) +
  geom_line() +
    facet_grid(.~Sex) +
  labs(title="Body Weight",
       x="Time on Diet (Weeks)",
       y="Change in Body Weight (g)")
```

# Lean Mass

```{r lean-mass-scatterplot, fig.cap="Scatter Plot of Lean Mass"}
exp.data %>%
  filter(assay=='Lean Mass') %>%
  ggplot(aes(y=Mass,x=diet.time,col=Diet)) +
  geom_point() +
  facet_grid(.~Sex) +
  stat_smooth() +
  labs(title="Lean Mass",
       x="Time on Diet (Weeks)",
       y="Lean Mass (g)")
```

```{r lean-mass-lineplot, fig.cap="Line Plot of Lean Mass"}
exp.data %>%
  filter(assay=='Lean Mass') %>%
  group_by(Diet,Sex,diet.time) %>%
  summarize(Average = mean(Mass, na.rm=T),
            SE = se(Mass)) %>%
  ggplot(aes(y=Average,
             ymin=Average-SE,
             ymax=Average+SE,
             x=diet.time,col=Diet)) +
  geom_errorbar(width=0.5) +
  geom_line() +
    facet_grid(.~Sex) +
  labs(title="Lean Mass",
       x="Time on Diet (Weeks)",
       y="Lean Mass (g)") 

exp.data %>%
  filter(assay=='Lean Mass') %>%
  group_by(Diet,Sex,diet.time) %>%
  summarize(Average = mean(Mass, na.rm=T),
            SE = se(Mass)) %>%
  ggplot(aes(y=Average,
             ymin=Average-SE,
             ymax=Average+SE,
             x=diet.time,col=Diet)) +
  geom_errorbar(width=0.5) +
  geom_line() +
    facet_grid(.~Sex) +
  labs(title="",
       x="Time on Diet (Weeks)",
       y="Lean Mass (g)") +
  scale_color_grey() +
  theme_classic() +
    theme(text=element_text(size=18),
          legend.position = c(0.75,0.75))
```

## Lean Mass Statistics

```{r lean-stats}
male.lme.lean.null <- lmer(Mass ~ diet.time + (1|animal.id), data=filter(exp.data, assay=='Lean Mass'&diet.time>=0&Sex=="M"))
male.lme.lean <- lmer(Mass ~ diet.time*Diet + (1|animal.id), data=filter(exp.data, assay=='Lean Mass'&diet.time>=0&Sex=="M"))
anova(male.lme.lean,male.lme.lean.null) %>% kable(caption="Male lean mass stats")

female.lme.lean.null <- lmer(Mass ~ diet.time + (1|animal.id), data=filter(exp.data, assay=='Lean Mass'&diet.time>=0&Sex=="F"))
female.lme.lean <- lmer(Mass ~ diet.time*Diet + (1|animal.id), data=filter(exp.data, assay=='Lean Mass'&diet.time>=0&Sex=="F"))
anova(female.lme.lean,female.lme.lean.null) %>% kable(caption="Female lean mass stats")

lmer(Mass ~ diet.time*Diet*Sex + (1|animal.id), data=filter(exp.data, assay=='Lean Mass'&diet.time>=0)) %>% summary %>% coef%>% kable(caption="Interaction of sex on lean mass")
```

```{r lean-mass-lineplot-change, fig.cap="Line Plot of Change in Lean Mass"}
exp.data %>%
  filter(assay=='Lean Mass') %>%
  group_by(Diet,Sex,diet.time) %>%
  summarize(Average = mean(Mass.Change, na.rm=T),
            SE = se(Mass.Change)) %>%
  ggplot(aes(y=Average,
             ymin=Average-SE,
             ymax=Average+SE,
             x=diet.time,col=Diet)) +
  geom_errorbar(width=0.5) +
  geom_line() +
    facet_grid(.~Sex) +
  labs(title="Lean Mass",
       x="Time on Diet (Weeks)",
       y="Change in Lean Mass (g)") 
```

# Fat Mass

```{r fat-mass-scatterplot, fig.cap="Scatter Plot of Fat Mass"}
exp.data %>%
  filter(assay=='Total Fat Mass') %>%
  ggplot(aes(y=Mass,x=diet.time,col=Diet)) +
  geom_point() +
  facet_grid(.~Sex) +
  stat_smooth() +
  labs(title="Fat Mass",
       x="Time on Diet (Weeks)",
       y="Fat Mass (g)")
```

```{r fat-mass-lineplot, fig.cap="Line Plot of Fat Mass"}
exp.data %>%
  filter(assay=='Total Fat Mass') %>%
  group_by(Diet,Sex,diet.time) %>%
  summarize(Average = mean(Mass,na.rm=T),
            SE = se(Mass)) %>%
  ggplot(aes(y=Average,
             ymin=Average-SE,
             ymax=Average+SE,
             x=diet.time,col=Diet)) +
  geom_errorbar(width=0.5) +
  geom_line() +
    facet_grid(.~Sex) +
  labs(title="Fat Mass",
       x="Time on Diet (Weeks)",
       y="Fat Mass (g)") 

exp.data %>%
  filter(assay=='Total Fat Mass') %>%
  group_by(Diet,Sex,diet.time) %>%
  summarize(Average = mean(Mass,na.rm=T),
            SE = se(Mass)) %>%
  ggplot(aes(y=Average,
             ymin=Average-SE,
             ymax=Average+SE,
             x=diet.time,col=Diet)) +
  geom_errorbar(width=0.5) +
  geom_line() +
    facet_grid(.~Sex) +
  labs(title="",
       x="Time on Diet (Weeks)",
       y="Fat Mass (g)") +
  scale_color_grey() +
  theme_classic() +
    theme(text=element_text(size=18),
          legend.position = c(0.75,0.75))
```


## Fat Mass Statistics

```{r fat-stats}
male.lme.weight.null <- lmer(Mass ~ diet.time + (1|animal.id), data=filter(exp.data, assay=='Total Fat Mass'&diet.time>=0&Sex=="M"))
male.lme.weight <- lmer(Mass ~ diet.time*Diet + (1|animal.id), data=filter(exp.data, assay=='Total Fat Mass'&diet.time>=0&Sex=="M"))
anova(male.lme.weight,male.lme.weight.null) %>% kable(caption="Male fat mass stats")

female.lme.weight.null <- lmer(Mass ~ diet.time + (1|animal.id), data=filter(exp.data, assay=='Total Fat Mass'&diet.time>=0&Sex=="F"))
female.lme.weight <- lmer(Mass ~ diet.time*Diet + (1|animal.id), data=filter(exp.data, assay=='Total Fat Mass'&diet.time>=0&Sex=="F"))
anova(female.lme.weight,female.lme.weight.null) %>% kable(caption="Female fat mass stats")

lmer(Mass ~ diet.time*Diet*Sex + (1|animal.id), data=filter(exp.data, assay=='Total Fat Mass'&diet.time>=0)) %>% summary %>% coef%>% kable(caption="Interaction of sex on fat mass")
```

```{r fat-mass-lineplot-change, fig.cap="Line Plot of Change in Fat Mass"}
exp.data %>%
  filter(assay=='Total Fat Mass') %>%
  group_by(Diet,Sex,diet.time) %>%
  summarize(Average = mean(Mass.Change,na.rm=T),
            SE = se(Mass.Change)) %>%
  ggplot(aes(y=Average,
             ymin=Average-SE,
             ymax=Average+SE,
             x=diet.time,col=Diet)) +
  geom_errorbar(width=0.5) +
  geom_line() +
    facet_grid(.~Sex) +
  labs(title="Fat Mass",
       x="Time on Diet (Weeks)",
       y="Change in Fat Mass (g)") 
```