---
title: "Quantification of AMPK Knockout Blots"
author: "Dave Bridges"
date: "June 18, 2020"
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: yes
    toc: yes
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=FALSE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

# Purpose

# Experimental Details

Blotted liver lysates for AMPK and ACC

# Raw Data


```{r data-input}
library(readxl) #loads the readr package

genotype.lanes <- c(rep("GFP",3), rep("Cre",3), rep("GFP",3), rep("Cre",4)) 
diet.lanes <- c(rep("control",6), rep("keto",7))

#this loads whatever the file is into a dataframe called exp.data if it exists
AMPK.filename <- 'Male AMPK.xls' #make this a separate line, you can use any variable you want
AMPK.data <- read_excel(AMPK.filename) %>% mutate(antibody="AMPK", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes))

acc.filename <- 'Male ACC.xls'
acc.data <- read_excel(acc.filename)  %>% mutate(antibody="ACC", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes))

chop.filename <- 'Male CHOP.xls'
chop.data <- read_excel(chop.filename)  %>% mutate(antibody="CHOP", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes))

fas.filename <- 'Male FAS.xls'
fas.data <- read_excel(fas.filename)  %>% mutate(antibody="FAS", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes))

pacc.filename <- 'Male pACC.xls'
pacc.data <- read_excel(pacc.filename)  %>% mutate(antibody="pACC", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes))

pS6.filename <- 'Male pS6.xls'
pS6.data <- read_excel(pS6.filename)  %>% mutate(antibody="pS6", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes))

S6.filename <- 'Male S6.xls'
S6.data <- read_excel(S6.filename)  %>% mutate(antibody="S6", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes))

totalprotein.filename <- 'Male total protein.xls'
totalprotein.data <- read_excel(totalprotein.filename)  %>% mutate(antibody="totalprotein", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes))

ACLY.filename <- 'Male ACLY quantification.xls'
ACLY.data <- read_excel(ACLY.filename)  %>% mutate(antibody="ACLY", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes))

pAMPK.filename <- 'Male pAMPK quantification.xls'
pAMPK.data <- read_excel(pAMPK.filename)  %>% mutate(antibody="pAMPK", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes))

mitoband1.filename <- 'Male mito top band quantification.xls'
mitoband1.data <- read_excel(mitoband1.filename)  %>% mutate(antibody="mitoband1", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes)) #this is the top band at 55 KD

mitoband2.filename <- 'Male mito second band quantification.xls'
mitoband2.data <- read_excel(mitoband2.filename)  %>% mutate(antibody="mitoband2", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes)) #this is the band at 48 KD

mitoband3.filename <- 'Male mito third band quantification.xls'
mitoband3.data <- read_excel(mitoband3.filename)  %>% mutate(antibody="mitoband3", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes)) #this is the band at 40 KD

mitoband4.filename <- 'Male mito fourth band quantification.xls'
mitoband4.data <- read_excel(mitoband4.filename)  %>% mutate(antibody="mitoband4", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes)) #this is the band at 30 KD

mitoband5.filename <- 'Male mito fifth band quantification.xls'
mitoband5.data <- read_excel(mitoband5.filename)  %>% mutate(antibody="mitoband5", Genotype=as.factor(genotype.lanes), Diet=as.factor(diet.lanes)) #this is the band at 20 KD

blot.data <-bind_rows(totalprotein.data, 
                      acc.data, 
                      pacc.data,
                      fas.data,
                      pS6.data,
                      S6.data,
                      chop.data,
                      pAMPK.data,
                      AMPK.data,
                      mitoband1.data,
                      mitoband2.data,
                      mitoband3.data,
                      mitoband4.data,
                      mitoband5.data)
```

These data can be found in **`r getwd()`** in files named **`r acc.filename`** and **`r pacc.filename`**.  This script was most recently updated on **`r date()`**.

# Analysis

```{r pACC-ACC}
normalized.data <-
  blot.data %>%
  select(Genotype,Diet, `Lane Name`, Signal, antibody) %>%
  pivot_wider(names_from=antibody, values_from=Signal) %>%
  mutate(Rel.ACC=ACC/totalprotein) %>%
  mutate(Norm.ACC=Rel.ACC/mean(Rel.ACC[Genotype=='GFP'&Diet=='control'])) %>%
  mutate(Rel.ACLY=ACLY/totalprotein) %>%
  mutate(Norm.ACLY=Rel.ACLY/mean(Rel.ACLY[Genotype=='GFP'&Diet=='control'])) %>%
  mutate(Rel.pACLY=pACLY/ACLY) %>%
  mutate(Norm.pACLY=Rel.pACLY/mean(Rel.pACLY[Genotype=='GFP'&Diet=='control'])) %>%
  mutate(Rel.AMPK=AMPK/totalprotein) %>%
  mutate(Norm.AMPK=Rel.AMPK/mean(Rel.AMPK[Genotype=='GFP'&Diet=='control'])) %>%
  mutate(Rel.CHOP=CHOP/totalprotein) %>%
  mutate(Norm.CHOP=Rel.CHOP/mean(Rel.CHOP[Genotype=='GFP'&Diet=='control'])) %>%
  mutate(Rel.FAS=FAS/totalprotein) %>%
  mutate(Norm.FAS=Rel.FAS/mean(Rel.FAS[Genotype=='GFP'&Diet=='control'])) %>%
  mutate(Rel.pACC=pACC/ACC) %>%
  mutate(Norm.pACC=Rel.pACC/mean(Rel.pACC[Genotype=='GFP'&Diet=='control'])) %>%
  mutate(Rel.S6=S6/totalprotein) %>%
  mutate(Norm.S6=Rel.S6/mean(Rel.S6[Genotype=='GFP'&Diet=='control'])) %>%
  mutate(Rel.pS6=pS6/S6) %>%
  mutate(Norm.pS6=Rel.pS6/mean(Rel.pS6[Genotype=='GFP'&Diet=='control'])) %>%
  mutate(Rel.pAMPK=pAMPK/AMPK) %>%
  mutate(Norm.pAMPK=Rel.pAMPK/mean(Rel.pAMPK[Genotype=='GFP'&Diet=='control']))
  mutate(Rel.mitoband1=mitoband1/totalprotein) %>%
  mutate(Norm.mitoband1=Rel.mitoband1/mean(Rel.mitoband1[Genotype=='GFP'&Diet=='control']))
  mutate(Rel.mitoband2=mitoband2/totalprotein) %>%
  mutate(Norm.mitoband2=Rel.mitoband2/mean(Rel.mitoband2[Genotype=='GFP'&Diet=='control']))
  mutate(Rel.mitoband3=mitoband3/totalprotein) %>%
  mutate(Norm.mitoband3=Rel.mitoband3/mean(Rel.mitoband3[Genotype=='GFP'&Diet=='control']))
  mutate(Rel.mitoband4=mitoband4/totalprotein) %>%
  mutate(Norm.mitoband4=Rel.mitoband4/mean(Rel.mitoband4[Genotype=='GFP'&Diet=='control']))
  mutate(Rel.mitoband5=mitoband5/totalprotein) %>%
  mutate(Norm.mitoband5=Rel.mitoband5/mean(Rel.mitoband5[Genotype=='GFP'&Diet=='control'])) #%>%
  #mutate(Rel.ACLY=ACLY/totalprotein) %>%
  #mutate(Norm.ACLY=Rel.ACLY/mean(Rel.ACLY[Genotype=='GFP'&Diet=='control']))
        
summary.data <-
  normalized.data %>%
  mutate(Genotype = relevel(Genotype, ref="GFP")) %>%
  group_by(Genotype,Diet) %>%
  summarize_at(.vars=vars(starts_with('Norm')), .funs=list(mean=mean,se=se,n=length))
  

 #once you get to this point Dave will help with adding the summary of means, se, length etc.
```

# Lipogenic Proteins

## Fatty Acid Synthase

```{r fas-barplot}
library(ggplot2)
ggplot(summary.data, aes(y=Norm.FAS_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.FAS_mean-Norm.FAS_se,
                    ymax=Norm.FAS_mean+Norm.FAS_se),
                width=0.5,
                position=position_dodge(width=0.75)) + 
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="FAS",
       title="Effects of AMPKa1/2 knockout on FAS levels")

library(broom)
aov(Rel.FAS ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for FAS levels, no interaction")

aov(Rel.FAS ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for FAS levels, with interaction")
```

## Acetyl-CoA Carboxylase

```{r acc-barplot}
ggplot(summary.data, aes(y=Norm.ACC_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.ACC_mean-Norm.ACC_se,
                    ymax=Norm.ACC_mean+Norm.ACC_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="ACC",
       title="Effects of AMPKa1/2 knockout on ACC levels")

aov(Rel.ACC ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for ACC levels, no interaction")

aov(Rel.ACC ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for ACC levels, with interaction")
```

```{r pACC-barplot}
ggplot(summary.data, aes(y=Norm.pACC_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.pACC_mean-Norm.pACC_se,
                    ymax=Norm.pACC_mean+Norm.pACC_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="pACC/ACC",
       title="Effects of AMPKa1/2 knockout on ACC phosphorylation")

aov(Rel.pACC ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for ACC phosphorylation, no interaction")

aov(Rel.pACC ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for ACC phosphorylation, with interaction")
```

## S6 Phosphorylation

```{r pS6-barplot}
ggplot(summary.data, aes(y=Norm.pS6_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.pS6_mean-Norm.pS6_se,
                    ymax=Norm.pS6_mean+Norm.pS6_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="pS6/S6",
       title="Effects of AMPKa1/2 knockout on S6 phosphorylation")

aov(Rel.pS6 ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for S6 phosphorylation, no interaction")

aov(Rel.pS6 ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for S6 phosphorylation, with interaction")
```

# Integrated Stress Response

```{r chop-barplot}
ggplot(summary.data, aes(y=Norm.CHOP_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.CHOP_mean-Norm.CHOP_se,
                    ymax=Norm.CHOP_mean+Norm.CHOP_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="CHOP",
       title="Effects of AMPKa1/2 knockout on CHOP levels")

aov(Rel.CHOP ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for CHOP levels, no interaction")

aov(Rel.CHOP ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for CHOP levels, with interaction")
```

## AMP Activated Protein Kinase 

```{r AMPK-barplot}
ggplot(summary.data, aes(y=Norm.AMPK_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.AMPK_mean-Norm.AMPK_se,
                    ymax=Norm.AMPK_mean+Norm.AMPK_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="AMPK",
       title="Effects of AMPKa1/2 knockout on AMPK levels")

aov(Rel.AMPK ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for AMPK levels, no interaction")

aov(Rel.AMPK ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for AMPK levels, with interaction")
```

```{r pAMPK-barplot}
ggplot(summary.data, aes(y=Norm.pAMPK_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.pAMPK_mean-Norm.pAMPK_se,
                    ymax=Norm.pAMPK_mean+Norm.pAMPK_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="pAMPK",
       title="Effects of AMPKa1/2 knockout on pAMPK levels")

aov(Rel.pAMPK ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for pAMPK levels, no interaction")

aov(Rel.pAMPK ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for pAMPK levels, with interaction")
```

## ATP citrate lyase

```{r ACLY-barplot}
ggplot(summary.data, aes(y=Norm.ACLY_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.ACLY_mean-Norm.ACLY_se,
                    ymax=Norm.ACLY_mean+Norm.ACLY_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="ACLY",
       title="Effects of AMPKa1/2 knockout on ACLY levels")

aov(Rel.ACLY ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for ACLY levels, no interaction")

aov(Rel.ACLY ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for ACLY levels, with interaction")
```

```{r pACLY-barplot}
ggplot(summary.data, aes(y=Norm.pACLY_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.pACLY_mean-Norm.pACLY_se,
                    ymax=Norm.pACLY_mean+Norm.pACLY_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="pACLY",
       title="Effects of AMPKa1/2 knockout on pACLY levels")

aov(Rel.pACLY ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for pACLY levels, no interaction")

aov(Rel.pACLY ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for pACLY levels, with interaction")
```

## Mitochondria Complexes Band 1

```{r mitoband1-barplot}
ggplot(summary.data, aes(y=Norm.mitoband1_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.mitoband1_mean-Norm.mitoband1_se,
                    ymax=Norm.mitoband1_mean+Norm.mitoband1_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="Mito band 1",
       title="Effects of AMPKa1/2 knockout on mitochondria complex band 1")

aov(Rel.mitoband1 ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for mitochondira band 1 levels, no interaction")

aov(Rel.mitoband1 ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for mitochondira band 1 levels, with interaction")
```

## Mitochondira Complexes band 2

```{r mitoband2-barplot}
ggplot(summary.data, aes(y=Norm.mitoband2_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.mitoband2_mean-Norm.mitoband2_se,
                    ymax=Norm.mitoband2_mean+Norm.mitoband2_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="Mito band 2",
       title="Effects of AMPKa1/2 knockout on mitochondria complex band 2")

aov(Rel.mitoband2 ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for mitochondira band 2 levels, no interaction")

aov(Rel.mitoband2 ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for mitochondira band 2 levels, with interaction")
```

## Mitochondria Complexes Band 3

```{r mitoband3-barplot}
ggplot(summary.data, aes(y=Norm.mitoband3_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.mitoband3_mean-Norm.mitoband3_se,
                    ymax=Norm.mitoband3_mean+Norm.mitoband3_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="Mito band 3",
       title="Effects of AMPKa1/2 knockout on mitochondria complex band 3")

aov(Rel.mitoband3 ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for mitochondira band 3 levels, no interaction")

aov(Rel.mitoband3 ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for mitochondira band 3 levels, with interaction")
```

## Mitochondira Complexes Band 4

```{r mitoband4-barplot}
ggplot(summary.data, aes(y=Norm.mitoband4_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.mitoband4_mean-Norm.mitoband4_se,
                    ymax=Norm.mitoband4_mean+Norm.mitoband4_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="Mito band 4",
       title="Effects of AMPKa1/2 knockout on mitochondria complex band 4")

aov(Rel.mitoband4 ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for mitochondira band 4 levels, no interaction")

aov(Rel.mitoband4 ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for mitochondira band 4 levels, with interaction")
```

## Mitochondria Complexes Band 5

```{r mitoband5-barplot}
ggplot(summary.data, aes(y=Norm.mitoband5_mean,x=Diet, fill=Genotype)) +
  geom_bar(stat='identity', position='dodge', width=0.75) +
  geom_errorbar(aes(ymin=Norm.mitoband5_mean-Norm.mitoband5_se,
                    ymax=Norm.mitoband5_mean+Norm.mitoband5_se),
                width=0.5,
                position=position_dodge(width=0.75)) +
  theme_classic() +
  theme(text=element_text(size=18)) +
  labs(y="Mito band 5",
       title="Effects of AMPKa1/2 knockout on mitochondria complex band 5")

aov(Rel.mitoband5 ~ Diet + Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for mitochondira band 5 levels, no interaction")

aov(Rel.mitoband5 ~ Diet * Genotype, data=normalized.data) %>%
  tidy %>%
  kable(caption="ANOVA for mitochondira band 5 levels, with interaction")
```

# Session Information

```{r session-information, echo=T}
sessionInfo()
```
