---
title: "QTL Analysis of Energy Expenditure from BXD Strains"
author: "Dave Bridges"
date: "February 11, 2022"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: yes
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

# Purpose

Use this file to generate bimbam files for GEMMA analyses

# Raw Data

Downloaded SNPs from GeneNetwork 

```{r data-input}
bxd.genotypes <- 'BXD-geno_8-9-17.txt'
library(readr)
bxd.genotype.data <- read_tsv(bxd.genotypes, skip=20,
                              col_types = cols(
  .default = col_character(),
  Chr = col_factor(levels=NULL),
  cM = col_double(),
  Mb = col_double()
)) %>%  
  mutate(`5XFAD`='B') %>% #strain 5x FAD is set to homozygous B allele
  mutate(`D2Gpnmb`='D') %>% #strain 5x FAD is set to homozygous B allele
  mutate(`D2`='D')  #strain 5x FAD is set to homozygous B allele

phenotype.data <-  'Strain Level Energy Expenditure Data.csv'
ee.phenotype.data <- 
  read_csv(phenotype.data) %>%
  filter(!is.na(MR_W))

```

These data can be found in **`r getwd()`**.  This script was most recently updated on **`r date()`**.

This analysis uses the BXD genotyping file at `r bxd.genotypes` which had `r dim(bxd.genotype.data)[1]` markers for `r dim(bxd.genotype.data)[2]` strains of mice.

## Making Energy Expenditure Phenotype Files

The phenotype file must be in the order of the genotypes with that length.  Missing phenotypes should be denoted with NA

```{r ee-phenotype-files}
strain.list <-
  bxd.genotype.data %>%
  mutate(`C57BL/6J`='B',
         `DBA/2J`='D') %>%
  select(ee.phenotype.data$Name) %>%
  colnames() %>%
  as.data.frame %>%
  rename("Strain"=1)

phenotype.table <-
  left_join(strain.list, ee.phenotype.data, by=c("Strain"="Name")) %>%
  select(Strain,MR_W, MR_adj) %>%
  arrange(Strain) 

phenotype.file <- 'ee_phenotype.csv'
phenotype.table %>% 
  select(MR_W) %>%
  write_csv(phenotype.file, col_names=F)


phenotype.file <- 'ee_adj_phenotype.csv'
phenotype.table %>% 
  select(MR_adj) %>%
  write_csv(phenotype.file, col_names=F)

ee.genotype.file <- 'bxd-ee-genotype.bimbam' 
bxd.genotype.data %>%
    mutate(`C57BL/6J`='B',
         `DBA/2J`='D') %>%
  select(-Chr,-cM) %>%
  mutate_at(.vars=vars(-("Locus")),
            .funs=funs(case_when(.=='B'~2,
                                            .=='D'~0))) %>%
  mutate(Major = 'B',
         Minor = 'D') %>%
  select(Locus,Major,Minor,strain.list$Strain) %>% 
  write_csv(ee.genotype.file, col_names=FALSE)

ee.phenotype.data[!(ee.phenotype.data$Name %in% phenotype.table$Strain),] %>%
  kable(caption="Strains without genotypes in genotype file")
```

Exported file with `r table(is.na(phenotype.table$Mean))['FALSE']` phenotypes.  This is based on `r dim(ee.phenotype.data)[1]` input phenotypes.  

# GEMMA Analysis

## Linear Mixed Models

```{sh bash-gemma-lmm, eval=T}
#generate relatedness matrix (uses standard matrix)
gemma -g bxd-ee-genotype.bimbam -p ee_phenotype.csv -gk 1 -o ee
#perform eigen decomposition
gemma -g bxd-ee-genotype.bimbam -p ee_phenotype.csv -k output/ee.cXX.txt -eigen -o ee
#run mixed linear model for fat.gain mass
gemma -g bxd-ee-genotype.bimbam -p ee_phenotype.csv -a snp-annotation.bimbam -k output/ee.cXX.txt -lmm 1 -o ee


#generate relatedness matrix (uses standard matrix)
gemma -g bxd-ee-genotype.bimbam -p ee_adj_phenotype.csv -gk 1 -o ee.adj
#perform eigen decomposition
gemma -g bxd-ee-genotype.bimbam -p ee_adj_phenotype.csv -k output/ee.adj.cXX.txt -eigen -o ee.adj
#run mixed linear model for fat.gain mass
gemma -g bxd-ee-genotype.bimbam -p ee_adj_phenotype.csv -a snp-annotation.bimbam -k output/ee.adj.cXX.txt -lmm 1 -o ee.adj
```

## BSLMM Analyses

```{sh bash-gemma-bslmm, eval=F, echo=F, message=F}
#removed 2 log from s and w
#run bslmm (see http://romainvilloutreix.alwaysdata.net/romainvilloutreix/wp-content/uploads/2017/01/gwas_gemma-2017-01-17.pdf), 250,000 burnin and 1,000,000 for number of sampling iterations
gemma -g bxd-ee-genotype.bimbam -p ee_adj_phenotype.csv -a snp-annotation.bimbam -bslmm 1 -w 250000 -s 1000000 -rpace 100 -wpace 1000 -maf 0.01 -o ee.adj

gemma -g bxd-ee-genotype.bimbam -p ee_adj_phenotype.csv -a snp-annotation.bimbam -bslmm 1 -w 250000 -s 1000000 -rpace 100 -wpace 1000 -maf 0.01 -o ee
```

## EE QTL Analysis

### Lean Mass Adjusted Energy Expenditure

```{r ee-adj-analysis}
ee.adj.lmm.filename <- 'output/ee.adj.assoc.txt'
ee.adj.lmm.results <- read.table(ee.adj.lmm.filename, sep='\t', header=T)
suggestive.pval <- 1E-5
gw.pval <- 5E-8

ee.adj.lmm.results %>%
  arrange(p_wald) %>% 
  filter(p_wald<gw.pval) %>%
  kable(caption="Genome-wide significant associations from mixed linear models for lean mass adjusted energy expenditure") 

ee.adj.lmm.results %>%
  arrange(p_wald) %>% 
  filter(p_wald<suggestive.pval) %>%
  kable(caption="Suggestive associations from mixed linear models for lean mass adjusted energy expenditure on NCD ") 

 ee.adj.lmm.results %>%
    arrange(p_wald) %>% 
    filter(p_wald<0.05) %>%
    head(10) %>%
    kable(caption="Top 10 nominal associations from mixed linear models for adjusted energy expenditure", digits=c(0,0,0,0,0,0,3,3,3,1,3,8)) 
 
 ee.adj.lmm.results %>%
   arrange(p_wald) %>% 
    group_by(chr) %>%
   head(50) %>%
  mutate(Lead.SNP = first(rs),
         lead.pval = min(p_wald)) %>%
  group_by(chr) %>%
  summarize(Region.Start=min(ps),
         Region.End = max(ps),
         Pval=min(lead.pval)) %>%
  mutate(Region.Size=Region.End-Region.Start) %>%
  arrange(Pval) %>%
  kable(caption="Top loci assocated with unadjusted energy expenditure")

library(qqman)
library(ggplot2) 

ee.adj.lmm.results %>%
  mutate(chr=as.numeric(as.character(chr))) %>%
  na.omit %>%
# Make the Manhattan plot on the gwasResults dataset
  manhattan(chr="chr", bp="ps", snp="rs", p="p_wald",col = color.scheme, 
            suggestiveline =-log10(suggestive.pval),
            genomewideline = -log10(gw.pval),
            cex.text=5) 
```

### Unadjusted Energy Expenditure

```{r ee-analysis}
ee.lmm.filename <- 'output/ee.assoc.txt'
ee.lmm.results <- read.table(ee.lmm.filename, sep='\t', header=T)


ee.lmm.results %>%
  arrange(p_wald) %>% 
  filter(p_wald<gw.pval) %>%
  kable(caption="Genome-wide significant associations from mixed linear models for unadjusted energy expenditure") 

ee.lmm.results %>%
  arrange(p_wald) %>% 
  filter(p_wald<suggestive.pval) %>%
  kable(caption="Suggestive associations from mixed linear models for unadjusted energy expenditure") 

ee.lmm.results %>%
   arrange(p_wald) %>% 
   head(10) %>%
   filter(p_wald<0.05) %>%
   kable(caption="Top 10 nominal associations from mixed linear models for unadjusted energy expenditure") 

ee.lmm.results %>%
   arrange(p_wald) %>% 
    group_by(chr) %>%
   head(50) %>%
  mutate(Lead.SNP = first(rs),
         lead.pval = min(p_wald)) %>%
  group_by(chr) %>%
  summarize(Region.Start=min(ps),
         Region.End = max(ps),
         Pval=min(lead.pval)) %>%
  mutate(Region.Size=Region.End-Region.Start) %>%
    arrange(Pval) %>%
  kable(caption="Top loci assocated with unadjusted energy expenditure")


ee.lmm.results %>%
  mutate(chr=as.numeric(as.character(chr))) %>%
  na.omit %>%
# Make the Manhattan plot on the gwasResults dataset
  manhattan(chr="chr", bp="ps", snp="rs", p="p_wald",col = color.scheme, 
            suggestiveline =-log10(suggestive.pval),
            genomewideline = -log10(gw.pval))
```

# Analysis of QTLs

```{ee-suggestive-snps}
sug.snps <- 
  ee.adj.lmm.results %>%
  arrange(p_wald) %>%
  distinct(chr, .keep_all=T) %>%
  filter(p_wald<suggestive.pval) %>%
  pull(rs)

sug.genotype.data <- bxd.genotype.data %>%
  filter(Locus %in% sug.snps) %>%
      mutate(`C57BL/6J`='B',
         `DBA/2J`='D') %>%
  select(Chr,Locus,ee.phenotype.data$Name)

snp.annotated.ee.adj.phenotype <- 
  sug.genotype.data %>% 
  pivot_longer(cols= -one_of('Chr','Locus','Name'), 
                                   names_to='Strain',
                                   values_to='Genotype') %>% 
  pivot_wider(names_from=c('Chr','Locus'),values_from='Genotype') %>%
  left_join(ee.phenotype.data,by =c("Strain"="Name"))
  

library(ggplot2)  
snp.annotated.ee.adj.phenotype %>% 
  group_by(`4_UNC8365581`) %>%
  summarize(ee.adj.Mean=mean(MR_adj,na.rm=T),
            ee.adj.SE=se(MR_adj)) %>%
  na.omit %>%
  ggplot(aes(y=ee.adj.Mean,
             x=`4_UNC8365581`,
             ymin=ee.adj.Mean-ee.adj.SE,
             ymax=ee.adj.Mean+ee.adj.SE)) +
  geom_bar(stat='identity') +
  geom_errorbar(width=0.5) +
  labs(title="Percent Fat Mass Gain",
       subtitle="Effect of UNC8365581 on Chr4",
       y="Adjusted Energy Expenditure (W)",
       x="Genotype") +
  theme(text=element_text(size=18))

snp.annotated.ee.adj.phenotype %>% 
  group_by(`14_rs4139535`) %>%
  summarize(ee.adj.Mean=mean(MR_adj,na.rm=T),
            ee.adj.SE=se(MR_adj)) %>%
  na.omit %>%
  ggplot(aes(y=ee.adj.Mean,
             x=`14_rs4139535`,
             ymin=ee.adj.Mean-ee.adj.SE,
             ymax=ee.adj.Mean+ee.adj.SE)) +
  geom_bar(stat='identity') +
  geom_errorbar(width=0.5) +
  labs(title="Percent Fat Mass Gain",
       subtitle="Effect of rs4139535 on Chr14",
       y="Adjusted Energy Expenditure (W)",
       x="Genotype") +
  theme_classic() +
  theme(text=element_text(size=18))

snp.annotated.ee.adj.phenotype %>% 
  group_by(`17_rs3701810`) %>%
  summarize(ee.adj.Mean=mean(MR_adj,na.rm=T),
            ee.adj.SE=se(MR_adj)) %>%
  na.omit %>%
  ggplot(aes(y=ee.adj.Mean,
             x=`17_rs3701810`,
             ymin=ee.adj.Mean-ee.adj.SE,
             ymax=ee.adj.Mean+ee.adj.SE)) +
  geom_bar(stat='identity') +
  geom_errorbar(width=0.5) +
  labs(title="Percent Fat Mass Gain",
       subtitle="Effect of rs3701810 on Chr17",
       y="Adjusted Energy Expenditure (W)",
       x="Genotype") +
  theme_classic() +
  theme(text=element_text(size=18))

snp.annotated.ee.adj.phenotype %>% 
  group_by(`X_rs48259763`) %>%
  summarize(ee.adj.Mean=mean(MR_adj,na.rm=T),
            ee.adj.SE=se(MR_adj)) %>%
  na.omit %>%
  ggplot(aes(y=ee.adj.Mean,
             x=`X_rs48259763`,
             ymin=ee.adj.Mean-ee.adj.SE,
             ymax=ee.adj.Mean+ee.adj.SE)) +
  geom_bar(stat='identity') +
  geom_errorbar(width=0.5) +
  labs(title="Percent Fat Mass Gain",
       subtitle="Effect of rs48259763 on ChrX",
       y="Adjusted Energy Expenditure (W)",
       x="Genotype") +
  theme_classic() +
  theme(text=element_text(size=18))
```

## BSLMM Analysis

### Adjusted for Lean Mass

```{r bslmm-adj-analysis}
ee.adj.bslmm.results.file <- 'output/ee.adj.param.txt'
ee.adj.bslmm.results <- read_tsv(ee.adj.bslmm.results.file, 
                               col_types = cols(chr=col_factor(levels=NULL))) %>%
                                 mutate(eff.size = beta*gamma) %>%
                                 arrange(-abs(eff.size))
  
ee.adj.bslmm.results %>% arrange(-gamma) %>% head(5) %>% kable(captin="Variants with the most likely sparse effects for lean mass adjusted energy expenditure.")

ee.adj.bslmm.hyp.file <- 'output/ee.adj.hyp.txt'
read_tsv(ee.adj.bslmm.hyp.file,
         col_types=cols(
           h = col_double(),
           pve = col_double(),
           rho = col_double(),
           pge = col_double(),
           pi = col_double(),
           n_gamma = col_integer()
)) %>%
  summarize_all(mean) %>%
  kable(caption="Mean hyperparameter estimates for lean mass adjusted energy expenditure.  Our snps explain some the variance (PVE), among these, some (n_gamma) SNPS explain a large amount amount of the variance explained by the SNPs (PGE).  Based on this, this phenotype is estimated to be about half of a BVSR model (rho)")
```

### Unadjusted for Lean Mass

```{r bslmm-ee-analysis}
ee.adj.bslmm.results.file <- 'output/ee.param.txt'
ee.adj.bslmm.results <- read_tsv(ee.adj.bslmm.results.file, 
                               col_types = cols(chr=col_factor(levels=NULL))) %>%
                                 mutate(eff.size = beta*gamma) %>%
                                 arrange(-abs(eff.size))
  
ee.adj.bslmm.results %>% arrange(-gamma) %>% head(5) %>% kable(captin="Variants with the most likely sparse effects for unadjusted analysis")

ee.adj.bslmm.hyp.file <- 'output/ee.hyp.txt'
read_tsv(ee.adj.bslmm.hyp.file,
         col_types=cols(
           h = col_double(),
           pve = col_double(),
           rho = col_double(),
           pge = col_double(),
           pi = col_double(),
           n_gamma = col_integer()
)) %>%
  summarize_all(mean) %>%
  kable(caption="Mean hyperparameter estimates for unadjusted analysis.  Our snps explain some the variance (PVE), among these, some (n_gamma) SNPS explain a large amount amount of the variance explained by the SNPs (PGE).  Based on this, this phenotype is estimated to be about half of a BVSR model (rho)")
```

# Session Information

```{r session-information, echo=T}
sessionInfo()
```
