---
title: "Analysis of Secreted Proteins from TSC Knockout Muscles"
author: "Dave Bridges"
date: "March 26, 2020"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: no
    toc: yes
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: yes
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figure/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')

library("knitcitations")
cleanbib()
options("citation_format" = "pandoc")
```

# Purpose

To broadly evaluate all potential myokines from mTORC1 activated muscles based on transcriptional changes from our RNAseq data 

# Experimental Details

Used RNAseq data compiled from our previous experiments  Used biomart to extract proteins with annotated signal peptides done via SignalP `r citep('10.1038/s41587-019-0036-z')`.

# Raw Data

```{r data-input}
library(readr) #loads the readr package
rnaseq.filename <- '../data/processed/Binary DESeq Results.csv' #make this a separate line, you can use any variable you want
rnaseq.analysed.filename <- '../data/processed/Binary Normalized Counts.csv'

#this loads whatever the file is into a dataframe called exp.data if it exists
rnaseq.stats <- read_csv(rnaseq.filename)
rnaseq.counts <- read_csv(rnaseq.analysed.filename)
```

## Annotated Proteins with Signal Peptides

```{r signalp-data}
library(biomaRt)

mouse.data <- useDataset('mmusculus_gene_ensembl', mart=useMart('ensembl'))
#listFilters(mouse.data) #to locate with_signalp
#listAttributes(mouse.data)  #to locate ensembl_gene_id

signalp.data <- 
  getBM(attributes=c('ensembl_gene_id','signalp'), 
      values = rnaseq.stats$Row.names, 
      mart = mouse.data) %>%
  mutate(signap = as.factor(signalp))

signalp.genes <-
  signalp.data %>%
  filter(signalp %in% c('SignalP-noTM')) %>%
  pull(ensembl_gene_id)
```

These data can be found in **`r getwd()`**.  The normalized RNAseq data can be found in a file named **`r rnaseq.filename`**.  This script was most recently updated on **`r date()`**.

# Analysis

The ENSEMBL dataset with genes annotated as having a signalP annotation includes `r length(signalp.genes)` or `r length(signalp.genes)/dim(signalp.data)[1]*100`% of all genes.

```{r potential-secreted-proteins}
rnaseq.secreted <-
  rnaseq.stats %>%
  filter(Row.names %in% signalp.genes) 

library(ggplot2)

sig.secreted <- 
  rnaseq.secreted %>%
  mutate(FC = 2^(log2FoldChange)) %>%
  filter(padj < 0.05,
         baseMean>2) %>%
  arrange(desc(abs(log2FoldChange))) 

sig.secreted %>%
  head(10) %>%
  kable(caption = "Top differentially expressed secreted proteins")
  
```


# Session Information

```{r session-information, echo=T}
sessionInfo()
```

# Bibliography

```{r bibliography, results='asis'}
write.bibtex(file="secreted-database-references.bib")
bibliography("markdown")
```