---
title: "Ketone Tolerance in Male DO Mice"
author: "Cody Cousineau and Dave Bridges"
date: "August 10, 2023"
output:
  html_document:
    toc: yes
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: no
    toc: yes
---



```{r setup, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

```{r data-input}
library(readr) #loads the readr package
filename <- "Diversity Outbred Data.csv" 

DO.Data <- read_csv(filename)


DO.baseline.ketone.data <- 
  DO.Data %>% 
  group_by(ID) %>%
  select(ID, BK_0, BK_15, BK_30, BK_45, BK_60)
 

DO.followup.ketone.data <- 
  DO.Data %>% 
  group_by(ID) %>%
  select(ID, FK_0, FK_15, FK_30, FK_45, FK_60)
  

DO.Data <- 
  DO.Data %>% 
  group_by(ID) %>%
  mutate(B_AUC = BK_0+BK_15+BK_30+BK_45+BK_60) %>%
  mutate(F_AUC = FK_0+FK_15+FK_30+FK_45+FK_60) %>%
  mutate(B_AUC.bl = B_AUC-(5*BK_0)) %>% #baseline corrected (time zero x number of time points)
  mutate(F_AUC.bl = F_AUC-(5*FK_0)) %>%
  mutate(AUC.change=F_AUC.bl-B_AUC.bl)
```

The raw data for this script can be found in `r filename`.  This script can be found in `r getwd()` and was most recently updated on `r date()`.

# Ketone levels

```{r ktt-dotplot, fig.cap="Ketone tolerance tests by treatments"}
DO.baseline.ktt.data.long <-
  pivot_longer(DO.baseline.ketone.data, cols = BK_0:BK_60, names_to='TimeStamp', values_to="ketone") %>%
  separate(TimeStamp, sep="BK_", into=c("Letter","Time")) %>%
  mutate(ketone = as.numeric(ketone)) %>%
  mutate(Time = as.numeric(Time))

library(ggplot2)
ggplot(DO.baseline.ktt.data.long,
            aes(y=ketone, x=Time)) +
  geom_jitter(width=0.3, alpha=0.5) +
  geom_smooth(method="loess") +
  labs(title="Ketone Tolerance Test",
       subtitle="Baseline KTT",
       y="Blood Ketones (mg/dL)") 

DO.followup.ktt.data.long <-
  pivot_longer(DO.followup.ketone.data, cols = FK_0:FK_60, names_to='TimeStamp', values_to="ketone") %>%
  separate(TimeStamp, sep="FK_", into=c("Letter","Time")) %>%
  mutate(ketone = as.numeric(ketone)) %>%
  mutate(Time = as.numeric(Time))

ggplot(DO.followup.ktt.data.long,
            aes(y=ketone, x=Time)) +
  geom_jitter(width=0.3, alpha=0.5) +
  geom_smooth(method="loess") +
  labs(title="Ketone Tolerance Test",
       subtitle="Follow-up KTT",
       y="Blood Ketones (mg/dL)") 

library(forcats)
DO.ktt.data.long <-
  DO.Data %>%
  select(ID, starts_with('BK'),starts_with('FK')) %>%
  pivot_longer(cols = BK_0:FK_60, names_to='TimeStamp', values_to="ketone") %>%
  separate(TimeStamp, sep="_", into=c("Letter","Time")) %>%
  mutate(ketone = as.numeric(ketone)) %>%
  mutate(Time = as.numeric(Time)) %>%
  mutate(Stage=fct_recode(Letter,
                          "Baseline"="BK",
                          "Follow-Up"="FK"))
  
DO.ktt.data.long %>%
  ggplot(aes(y=ketone,x=Time,col=Stage)) +
    geom_jitter(width=0.3, alpha=0.5) +
  geom_smooth(method="loess") +
  labs(title="Diversity Outbred Mice",
       y="Blood Ketone Bodies (mg/dL)") +
  theme_classic(base_size=16) +
  theme(legend.position=c(0.75,0.75)) +
  scale_color_grey()
```

# Area Under the Curve

```{r AUC-analysis}
DO.all.AUC.data <-
  DO.Data %>%
  group_by(ID) %>%
  select(ID, B_AUC, F_AUC,B_AUC.bl, F_AUC.bl) %>%
  mutate(AUC.change=F_AUC.bl-B_AUC.bl)

ggplot(DO.all.AUC.data,
            aes(y=F_AUC, x=B_AUC)) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(title="AUC of Ketone Tolerance Tests",
       subtitle="Comparison of AUCs",
       y="Follow-Up AUC",
       x="Baseline AUC") +
    geom_abline(slope=1,lty=2) +
  expand_limits(x = 0, y = 0)

ggplot(DO.all.AUC.data %>% na.omit,
            aes(y=F_AUC.bl, x=B_AUC.bl)) +
  geom_point() +
  geom_smooth(method="lm", show.legend = T) +
  labs(title="AUC of Ketone Tolerance Tests",
       subtitle="Baseline Corrected AUC",
       y="Follow-Up AUC",
       x="Baseline AUC") +
  geom_abline(slope=1,show.legend = T,lty=2) +
  expand_limits(x = 0, y = 0) 

DO.all.AUC.data %>%
  select(ID,ends_with('.bl'),AUC.change) %>%
  na.omit %>%
  pivot_longer(ends_with('.bl'), values_to="AUC") %>%
  mutate(Stage=fct_recode(name,
                       "Baseline"="B_AUC.bl",
                       "Follow-Up"="F_AUC.bl")) %>%
  ggplot(aes(y=AUC,x=Stage,group=ID,col=AUC.change>0)) +
  geom_line() +
  labs(y="Baseline adjusted area under curve",
       x="",
       title="Changes in BHB disposal") +
  theme_classic(base_size=16) +
  theme(legend.position=c(0.2,0.8))+
  scale_color_grey(name="Increased with Diet")

library(ggrepel)
DO.all.AUC.data %>%
  select(ID,ends_with('.bl'),AUC.change) %>%
  na.omit %>%
  pivot_longer(ends_with('.bl'), values_to="AUC") %>%
  mutate(Stage=fct_recode(name,
                       "Baseline"="B_AUC.bl",
                       "Follow-Up"="F_AUC.bl")) %>%
  ggplot(aes(y=AUC,x=Stage,group=ID,col=AUC.change>0)) +
  geom_line() +
  geom_label_repel(aes(label=ID),size=2) 
```


```{r KTT-associations}

ggplot(DO.Data,
            aes(y=B_AUC, x=BK_0)) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(title="AUC of Ketone Tolerance Tests VS Time Zero Ketones",
       subtitle="Comparison of Baseline AUCs and Baseline T0 Ketones",
       y="Baseline AUC",
       x="Baseline Ketones") +
    geom_abline(slope=1,lty=2) +
  expand_limits(x = 0, y = 0)

ggplot(DO.Data,
            aes(y=B_AUC, x=FK_0)) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(title="AUC of Ketone Tolerance Tests VS Time Zero Ketones",
       subtitle="Comparison of Baseline AUCs and Follow-Up T0 Ketones",
       y="Baseline AUC",
       x="Follow-Up Ketones") +
    geom_abline(slope=1,lty=2) +
  expand_limits(x = 0, y = 0)

ggplot(DO.Data,
            aes(y=F_AUC, x=BK_0)) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(title="AUC of Ketone Tolerance Tests VS Time Zero Ketones",
       subtitle="Comparison of Follow-Up AUCs and Baseline T0 Ketones",
       y="Follow-Up AUC",
       x="Baseline Ketones") +
    geom_abline(slope=1,lty=2) +
  expand_limits(x = 0, y = 0)

ggplot(DO.Data,
            aes(y=F_AUC, x=FK_0)) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(title="AUC of Ketone Tolerance Tests VS Time Zero Ketones",
       subtitle="Comparison of Follow-Up AUCs and Follow-Up T0 Ketones",
       y="Follow-Up AUC",
       x="Follow-Up Ketones") +
    geom_abline(slope=1,lty=2) +
  expand_limits(x = 0, y = 0)

DO.all.AUC.data <-
  DO.Data %>%
  group_by(ID) %>%
  select(ID, B_AUC, F_AUC,B_AUC.bl, F_AUC.bl, BK_0, FK_0) %>%
  mutate(AUC.change=F_AUC.bl-B_AUC.bl)

ggplot(DO.all.AUC.data,
            aes(y=B_AUC.bl, x=BK_0)) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(title="AUC of Ketone Tolerance Tests VS Time Zero Ketones",
       subtitle="Comparison of Baseline AUCs and Baseline T0 Ketones",
       y="Baseline AUC.bl",
       x="Baseline Ketones") +
    geom_abline(slope=1,lty=2) +
  expand_limits(x = 0, y = 0)

ggplot(DO.Data,
            aes(y=B_AUC.bl, x=FK_0)) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(title="AUC of Ketone Tolerance Tests VS Time Zero Ketones",
       subtitle="Comparison of Baseline AUCs and Follow-Up T0 Ketones",
       y="Baseline AUC.bl",
       x="Follow-Up Ketones") +
    geom_abline(slope=1,lty=2) +
  expand_limits(x = 0, y = 0)

ggplot(DO.Data,
            aes(y=F_AUC.bl, x=BK_0)) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(title="AUC of Ketone Tolerance Tests VS Time Zero Ketones",
       subtitle="Comparison of Follow-Up AUCs and Baseline T0 Ketones",
       y="Follow-Up AUC.bl",
       x="Baseline Ketones") +
    geom_abline(slope=1,lty=2) +
  expand_limits(x = 0, y = 0)

ggplot(DO.Data,
            aes(y=F_AUC.bl, x=FK_0)) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(title="AUC of Ketone Tolerance Tests VS Time Zero Ketones",
       subtitle="Comparison of Follow-Up AUCs and Follow-Up T0 Ketones",
       y="Follow-Up AUC.bl",
       x="Follow-Up Ketones") +
    geom_abline(slope=1,lty=2) +
  expand_limits(x = 0, y = 0)

```

# Statistics

At a population level is there an increase or decrease in AUC

```{r auc-comparasion-stats}
library(broom)
t.test(DO.all.AUC.data$AUC.change,mu=0) %>% tidy %>% kable(caption="One sample t test for whether the change in AOC is greater than zero", digits=c(2,2,8,0,2,2,0,0))
  
with(DO.all.AUC.data, wilcox.test(B_AUC,F_AUC,paired=T))%>% tidy %>% kable(caption="Pairwise wilcoxon test comparing baseline to follow up AUC", digits=c(0,8,0,0))
  
with(DO.all.AUC.data, wilcox.test(B_AUC.bl,F_AUC.bl,paired=T))%>% tidy %>% kable(caption="Pairwise wilcoxon test comparing corrected baseline to follow up AUC", digits=c(0,8,0,0))

```

How many strains have an increase vs decrease in AUC

```{r strain-count}
DO.all.AUC.data %>%
  group_by(AUC.change<0) %>%
  count -> strain.count

strain.count %>%
  kable(caption="Number of strains with a decrease in corrected AUC from baseline to follow up.")

strain.count %>% filter(!is.na(`AUC.change < 0`)) %>% pull(n) %>%
  chisq.test(p=c(1,1),rescale.p = T) %>%
  tidy %>%
  kable(caption='Chi-squared test for equal increasers and decreasers in baseline corrected AUC',
        digits=c(1,10,1))

DO.all.AUC.data %>%
  arrange(AUC.change) %>%
  kable(caption="Strains sorted by corrected change in AUC")
```



## Is there a relationship between baseline and follow-up AUC?

```{r auc-stats}
library(car)
leveneTest(DO.Data$B_AUC.bl,DO.Data$F_AUC.bl) %>% 
  tidy %>% 
  kable(caption="Levene's test comparing baseline to follow up baseline adjusted KTT AUC",
        digits=c(2,99,1,1))

DO.Data %>%
  ungroup %>%
  rename("AUC.Baseline"="B_AUC",
         "AUC.Followup"="F_AUC",
         "AUC.Baseline.Cor"="B_AUC.bl",
         "AUC.Followup.Cor"="F_AUC.bl") %>%
  summarize_at(.vars=vars('AUC.Baseline':'AUC.change'),
               .funs=list(mean=~mean(.,na.rm=T),
                          sd=~sd(.,na.rm=T),
                          n=~length(.),
                          shapiro=~shapiro.test(.)$p.value)) %>%
  pivot_longer(cols=everything(),
               names_sep="_",
               names_to=c("Measure","Statistic")) %>%
  pivot_wider(names_from=Statistic,
              values_from=value) -> summary.statistics

summary.statistics %>%
  kable(caption="Summary Statistics for DO KTT")

library(broom)
 with(DO.all.AUC.data, cor.test(B_AUC,F_AUC,method="spearman")) %>% tidy %>% kable(caption="Correlation between baseline and follow up AUC")
 
  with(DO.Data, cor.test(B_AUC.bl,F_AUC.bl,method="spearman")) %>% tidy %>% kable(caption="Correlation between corrected baseline and follow up AUC")
  
  with(DO.Data, cor.test(B_AUC,BK_0,method="spearman")) %>% tidy %>% kable(caption="Correlation between baseline AUC and baseline ketones")
  
  with(DO.Data, cor.test(B_AUC,FK_0,method="spearman")) %>% tidy %>% kable(caption="Correlation between baseline AUC and follow-up ketones")
  
  with(DO.Data, cor.test(F_AUC,BK_0,method="spearman")) %>% tidy %>% kable(caption="Correlation between follow-up AUC and baseline ketones")
    
  with(DO.Data, cor.test(F_AUC,FK_0,method="spearman")) %>% tidy %>% kable(caption="Correlation between follow-up AUC and follow-up ketones")
  
  with(DO.all.AUC.data, cor.test(B_AUC.bl,BK_0,method="spearman")) %>% tidy %>% kable(caption="Correlation between baseline AUC and baseline ketones")
  
  with(DO.all.AUC.data, cor.test(B_AUC.bl,FK_0,method="spearman")) %>% tidy %>% kable(caption="Correlation between baseline AUC and follow-up ketones")
  
  with(DO.all.AUC.data, cor.test(F_AUC.bl,BK_0,method="spearman")) %>% tidy %>% kable(caption="Correlation between follow-up AUC and baseline ketones")
    
  with(DO.all.AUC.data, cor.test(F_AUC.bl,FK_0,method="spearman")) %>% tidy %>% kable(caption="Correlation between follow-up AUC and follow-up ketones")
```

The ratio of $SD_{followup}/SD_{baseline}$ was `r filter(summary.statistics, Measure %in% c('AUC.Baseline.Cor','AUC.Followup.Cor'))[2,'sd']/filter(summary.statistics, Measure %in% c('AUC.Baseline.Cor','AUC.Followup.Cor'))[1,'sd']`

## Bayesian Approach to Baseline vs Followup

First I wanted to evaluate the relationship between baseline and followup KTT (area under the curve, baseline corrected) in the DO mice.  I used default priors with a normal (gausian) distribution

```{r baseline-followup-bayesian}
#| output: false

library(brms)

auc.fit.brms <- brm(F_AUC.bl~B_AUC.bl,
                    data=DO.all.AUC.data,
                    sample_prior = TRUE)
```

Here is a summary of this modeling

```{r baseline-followup-bayesian-analysis}
prior_summary(auc.fit.brms) %>% kable(caption="Prior probabilities summary for relationship betwen baseline and followup AUC")
kable(data.frame(
  Parameter = names(rhat(auc.fit.brms)),
  Rhat = format(rhat(auc.fit.brms), nsmall = 5)),
caption="Rhat values for model fitting between baseline and followup (should b between 0.99 and 1.01 for convergence",
row.names = F)

library(broom.mixed)
tidy(auc.fit.brms, effects = "ran_pars", ess = TRUE) %>% kable(caption="Model random effects for the relationship between baseline and followup AUC.  ESS should be >400 samples")

library(broom.mixed)
tidy(auc.fit.brms) %>% kable(caption="Summary of the relationship between baseline and followup")
plot(auc.fit.brms)
hypothesis(auc.fit.brms, "B_AUC.bl>0") # testing for whether baseline AUC has any effect on followup AUC

as_draws_df(auc.fit.brms) %>%
  ggplot(aes(x=b_B_AUC.bl)) +
  geom_density(fill=color.scheme[1]) +
  geom_vline(xintercept=0,color=color.scheme[2],lty=2) +
  labs(y="",
       x="Effect of baseline AUC on followup AUC",
       title="Posterior probabilities") +
  theme_classic(base_size=16)
kable(bayes_R2(auc.fit.brms),caption="Estimates for R2 between baseline and followup AUC")


r2.probs <- bayes_R2(auc.fit.brms, summary=F) #summary false is to get the posterior probabilities
ggplot(data=r2.probs, aes(x=R2)) +
  geom_density(fill=color.scheme[1]) +
  geom_vline(xintercept=0,color=color.scheme[2],lty=2) +
  labs(y="",
       x="R2",
       title="Posterior probabilities for R2") +
  lims(x=c(0,1)) +
  theme_classic(base_size=16)
```

# Association with Cholesterol

```{r ktt-cholesterol}
# are these values normally distributed?
DO.Data %>%
  ungroup %>%
  select('B_AUC.bl','F_AUC.bl','AUC.change') %>%
  summarize_all(.funs=list(shapiro=~shapiro.test(.)$p.value)) %>%
  kable(caption="Shapiro.wilk tests for each KTT parameter")
# can use parametric values                                                                   

library(tibble)
DO.Data %>%
  mutate(Weight.Gain=BW6-BW3) %>%
  select(ID,B_AUC.bl,F_AUC.bl,AUC.change,BW3,Weight.Gain,Cholesterol_Baseline,Cholesterol_Followup,TG_Baseline,	TG_Followup,BG1,BG2) %>%
  mutate("Change in Triglycerides"=TG_Followup-TG_Baseline,
         "Change in Cholesterol"=Cholesterol_Followup-Cholesterol_Baseline,
         "Change in Glucose"=BG2-BG1) %>%
  rename("Baseline KTT"="B_AUC.bl",
         "Followup KTT"="F_AUC.bl",
         "Change in KTT"="AUC.change",
         "Baseline Weight"="BW3",
         "Change in Weight"="Weight.Gain",
         "Baseline Cholesterol"="Cholesterol_Baseline",
         "Followup Cholesterol"="Cholesterol_Followup",
         "Baseline Triglycerides"="TG_Baseline",
         "Folowup Triglycerides"="TG_Followup",
         "Baseline Glucose"="BG1",
         "Followup Glucose"="BG2") -> KTT.data.all

KTT.matrix <- 
  KTT.data.all %>%
  na.omit %>%
  column_to_rownames(var="ID") 
  

library(corrplot)
cor.matrix <- cor(KTT.matrix,use="all.obs")[1:3,4:14]
corrplot(cor.matrix,method="square",tl.col = "black")  
corrplot(cor.matrix,method="shade",tl.col = "black")  
corrplot(cor.matrix,method="number",tl.col = "black")  

ggplot(KTT.data.all,
       aes(y=`Change in Weight`,
           x=`Followup KTT`)) +
  geom_point() +
  geom_smooth(method="lm") +
  theme_classic(base_size = 16)

ggplot(KTT.data.all,
       aes(y=`Change in Cholesterol`,
           x=`Baseline KTT`)) +
  geom_point() +
  geom_smooth(method="lm") +
  theme_classic(base_size = 16)-> plot.cor.1

ggplot(KTT.data.all,
       aes(y=`Change in Weight`,
           x=`Followup KTT`)) +
  geom_point() +
  geom_smooth(method="lm") +
  theme_classic(base_size = 16)-> plot.cor.2

ggplot(KTT.data.all,
       aes(y=`Change in Weight`,
           x=`Change in KTT`)) +
  geom_point() +
  geom_smooth(method="lm") +
  theme_classic(base_size = 16) -> plot.cor.3

plot.cor.1
plot.cor.2
plot.cor.3

library(gridExtra)
grid.arrange(plot.cor.1,plot.cor.2,plot.cor.3,nrow=1)

library(rstatix)
cor_pmat(KTT.matrix)[1:3,4:15] %>% as.data.frame -> cor.pvals
rownames(cor.pvals) <- c("Baseline KTT","Followup KTT","Change in KTT")
cor.pvals %>% kable(caption="P-values for correlation tests")


```

## Regression Tree for Cholesterol Changes

```{r regression-tree-cholesterol}
library(rpart)

prepared_data <- DO.Data %>%
  mutate(`Change in Cholesterol` = as.numeric(Cholesterol_Followup) - as.numeric(Cholesterol_Baseline)) %>%
    select(-ID,-Cholesterol_Followup) 

tree.chol <- rpart(`Change in Cholesterol` ~ ., data = prepared_data, method = "anova")

library(rpart.plot)
rpart.plot(tree.chol, uniform = TRUE, main = "Regression Tree for Cholesterol Changes", 
        extra = 1, branch = 0.5, digits = 2, fallen.leaves = TRUE)
```

### Bootstrapping of regression trees

```{r regression-tree-cholesterol-bootstrapping}
#used perplexity.ai
bootstrap_trees <- function(data, n_bootstrap = 10000) {
  results <- list()
  for (i in 1:n_bootstrap) {
    boot_sample <- data[sample(nrow(prepared_data), replace = TRUE), ]
    boot_tree <- rpart(`Change in Cholesterol` ~ ., data = data, method = "anova")
    results[[i]] <- boot_tree
  }
  return(results)
}

bootstrapped_trees <- bootstrap_trees(prepared_data)

#analyse node frequency
node_frequency <- function(tree_list) {
  node_counts <- table(unlist(lapply(tree_list, function(t) t$frame$var)))
  node_freq <- prop.table(node_counts)
  return(sort(node_freq, decreasing = TRUE))
}

node_freq <- node_frequency(bootstrapped_trees)
print(node_freq)

#barplot of node frequencies
ggplot(as.data.frame(node_freq) %>% mutate(Var1=fct_recode(as.factor(Var1),"None"="<leaf>")), 
       aes(x = reorder(Var1, Freq), y = Freq)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Node Frequencies from Bootstrapped Trees", y = "Frequency") +
  theme_classic(base_size = 16) 
```

### Plotting Glycerol Followup vs Change in Cholesterol

```{r glycerol-followup-cholesterol}
library(ggplot2)
ggplot(prepared_data,
       aes(y=`Change in Cholesterol`,
           x=Glycerol_Followup)) +
  geom_point() +
  geom_smooth(method="lm") +
  theme_classic(base_size = 16) 

lm(`Change in Cholesterol`~Glycerol_Followup,data=prepared_data) %>%
  tidy %>%
  kable(caption="Linear model for change in cholesterol vs followup glycerol")
```

### Plotting Ketones During KTT vs Change in Cholesterol

```{r fk45-followup-cholesterol}
library(ggplot2)
ggplot(prepared_data,
       aes(y=`Change in Cholesterol`,
           x=FK_45)) +
  geom_point() +
  geom_smooth(method="lm") +
  theme_classic(base_size = 16) 

lm(`Change in Cholesterol`~FK_45,data=prepared_data) %>%
  tidy %>%
  kable(caption="Linear model for change in cholesterol vs followup glycerol")
```

### Bayesian Analysis of Node Frequencies

```{r bayesian-node-frequencies}
library(brms)
library(dplyr)
library(kableExtra)

# 1. Your observed frequency of <leaf>
leaf_frequency <- 0.571  # From your previous rpart analysis
n_trees <- 1000  # Number of bootstrapped trees (adjust if different)
leaf_count <- round(leaf_frequency * n_trees) #Number of leaf nodes

# 2. Create the data for brms
data_for_brms <- data.frame(
  successes = leaf_count,
  trials = n_trees
)

# 3. Define and fit the brms model
#    - Beta prior: Here, Beta(1, 1) is a uniform prior (all proportions equally likely a priori).
#      You can change this to a more informative prior if you have prior knowledge.

proportion_model <- brm(
  successes | trials(trials) ~ 1,  # Intercept-only model (modeling a single proportion)
  data = data_for_brms,
  family = binomial(),           # Binomial family (for proportions)
  prior = prior(beta(1, 1), class = "Intercept"), #Set the beta prior 
  sample_prior = TRUE,          # draw samples from the prior for prior predictive checks
  chains = 4,                  # Multiple chains for MCMC convergence
  iter = 2000,                  # Number of MCMC iterations per chain
  warmup = 1000                 # Warmup iterations
)

# Summarize the model
summary(proportion_model)

# 4. Calculate the Bayes Factor for proportion > 0
#    - We'll approximate this by calculating the posterior probability
#      that the proportion is greater than 0.  Since a proportion *must* be
#      greater than zero, technically we are calculating the proportion the 
#      the value is greater than 0 *given your prior*.  But since the prior
#      assigns some probability to values less than 0 (which is not possible),
#      this test essentially calculates the probability that the proportion
#      is very small.

hypothesis_result <- hypothesis(proportion_model, "Intercept > 0")
print(hypothesis_result)
kable(bayes_R2(proportion_model),caption="Estimates for R2 between baseline and followup AUC")
plot(proportion_model)
pp_check(proportion_model)

prior_summary(proportion_model) %>% kable(caption="Prior probabilities summary for relationship betwen baseline and followup AUC")
kable(data.frame(
  Parameter = names(rhat(proportion_model)),
  Rhat = format(rhat(proportion_model), nsmall = 5)),
  caption="Rhat values for model fitting between baseline and followup (should b between 0.99 and 1.01 for convergence",
  row.names = F)

```

# Session Information

```{r session-info}
sessionInfo()
```

